---
title: "Take-home Ex02"
author: "Jiang Yuxi"
format: html
editor: visual
date: "May 22, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Exploring Vast Challenge 2024 - Mini-challenge 1

# 1. Overview

## 1.1 Setting the scene

## 1.2 Key Objectives of this assignment

# 2. Data Import & Processing

## 2.1 Loading Relevant Packages and Data

| Packages Used | Purpose |
|----------------------------------|--------------------------------------|
| jsonlite, sf | Importing JSON file and geojson for geographical data |
| tidyverse, lubridate, dplyr, readr, reshape2, data.table | Data wrangling and reshaping in preparation for visualisation and wrap text function |
| ggplot2, tidygraph, patchwork, sf, ggrepel | Statistical graph and geographical plots with points and polygons and tidying of graphs for clarity and aesthetics |
| DT, igraph, ggraph, plotly, visNetwork, ggiraph | For interactive graph and data table for user to drill down on area of interest |

Load packages

```{r}
pacman::p_load(tidyverse, jsonlite, tidygraph, ggraph, lubridate)
```

Read JSON file

```{r}
mc1_data <- fromJSON ("VAST_Challenge_2026_MC1/MC1_final_00.json")
```

Save parsed object for quick reloads

```{r}
write_rds(mc1_data, "data/mc1_data.rds")
```

## 2.2 Inspect Data Structure

to see which fields are nested and which are flat

```{r}
glimpse(mc1_data)
```

## 2.3 Flat Nested Columns

separated the dataset into simple columns (flat, easy to analyze) and nested list columns (require flattening).

mutate() converts strings like "2046-05-17T09:00:00" into proper timestamps.

parse_number() strips \$ and % so you can plot stock price and percent change.

This table is perfect for time-series plots, sentiment trends, and event timelines.

```{r}
# Convert rounds into tibble
df_flat <- as_tibble(mc1_data$rounds)
glimpse(df_flat)
```

```{r}
# --- Step 1: Flatten environment_context ---
df_flat <- df_flat %>%
  unnest_wider(environment_context) %>%                 # expand 10 subfields
  unnest_wider(market_snapshot, names_sep = "_")        # expand nested market_snapshot

# --- Step 2: Round-level context (macro trends) ---
context_df <- df_flat %>%
  select(hour, event_headline, event_narrative,
         social_state, market_snapshot_stock_price,
         market_snapshot_percent_change, market_snapshot_sentiment)

# --- Step 3: Communications (micro actions) ---
comm_df <- df_flat %>%
  select(hour, communications) %>%
  unnest(communications)   # each message becomes a row

# --- Step 4: Participants ---
participants_df <- df_flat %>%
  select(hour, participants) %>%
  unnest(participants)     # each participant becomes a row

# --- Step 5: Alerts ---
alerts_df <- df_flat %>%
  select(hour, social_manager_alerts) %>%
  unnest_longer(social_manager_alerts)

# --- Step 6: Deadlines ---
deadlines_df <- df_flat %>%
  select(hour, critical_deadlines) %>%
  unnest_longer(critical_deadlines)

# --- Step 7: News ---
news_df <- df_flat %>%
  select(hour, news) %>%
  unnest_longer(news)

# --- Step 8: Hashtags ---
hashtags_df <- df_flat %>%
  select(hour, market_snapshot_trending_hashtags) %>%
  unnest_longer(market_snapshot_trending_hashtags)

# --- Step 9: Media Events ---
media_df <- df_flat %>%
  select(hour, media_events) %>%
  unnest_longer(media_events)

# --- Step 10: External Actor Actions ---
actors_df <- df_flat %>%
  select(hour, external_actor_actions) %>%
  unnest_longer(external_actor_actions)

# --- Step 11: Agents Unavailable ---
agents_df <- df_flat %>%
  select(hour, agents_unavailable) %>%
  unnest_longer(agents_unavailable)

# --- Step 12: Join everything back on hour ---
combined_df <- context_df %>%
  left_join(comm_df, by = "hour") %>%
  left_join(participants_df, by = c("hour", "agent_id")) %>%
  left_join(alerts_df, by = "hour") %>%
  left_join(deadlines_df, by = "hour") %>%
  left_join(news_df, by = "hour") %>%
  left_join(hashtags_df, by = "hour") %>%
  left_join(media_df, by = "hour") %>%
  left_join(actors_df, by = "hour") %>%
  left_join(agents_df, by = "hour")

 # --- Step 13: unnest nested columns---
combined_df <- combined_df %>%
  unnest_wider(internal_state, names_sep = "_") %>%
  unnest_wider(agent_round_metadata, names_sep = "_")

# Inspect the combined dataset
glimpse(combined_df)
```

**Structure of the Data Round context (macro level)**

-   hour: timestamp of the round (e.g., "2046-05-17T09:00:00")

-   event_headline: short summary of the round’s main event

-   event_narrative: longer description of what happened

-   social_state: public sentiment or chatter at that time

-   market_snapshot_stock_price, market_snapshot_percent_change, market_snapshot_sentiment: financial indicators

**Message details (micro level)**

-   message_id: unique ID for each communication

-   channel: communication channel (e.g., comms_huddle)

-   recipients: who received the message (often "ALL")

-   message_type: type of communication (broadcast, direct, etc.)

-   responding_to: links replies to earlier messages

-   content: the actual text of the message

-   timestamp: precise time within the round

**Agent information**

-   agent_id: unique identifier (e.g., legal_agent, quality_agent)

-   agent_role.x / agent_role.y: role categories (legal, platform_trust, PR, social_media)

-   agent_label.x / agent_label.y: human‑readable labels (Legal‑Agent, PR‑Agent, etc.)

-   declared_action: explicit actions taken (often NA if none declared)

**Internal state (flattened)**

-   internal_state_reacting, internal_state_rationalizing, internal_state_deliberating: cognitive/emotional notes about the agent’s reasoning during that message.

**Agent round metadata**

-   agent_round_metadata_sentiment_at_turn: sentiment snapshot for that agent at that turn

-   agent_round_metadata_action_classification: classification of the agent’s action

**Additional context signals**

-   social_manager_alerts, critical_deadlines, news, market_snapshot_trending_hashtags, media_events, external_actor_actions, agents_unavailable: contextual signals, often sparse (many NA or empty lists).

```{r}
saveRDS(combined_df, "data/combined_df_raw.rds")
```

## 2.4 Data Wrangling

### 2.4.1 Data Type Conversion

Convert stock price: Strip the \$ sign and convert to numeric. 
Convert percent change: Remove % and convert to numeric. 
Standardize timestamps: Convert hour and timestamp to proper datetime objects.

```{r}
combined_df <- readRDS("data/combined_df_raw.rds")
names(combined_df)
```

```{r}
cleaned_df <- combined_df %>%
  mutate(
    market_snapshot_stock_price = as.numeric(gsub("[$]", "", market_snapshot_stock_price)),
    market_snapshot_percent_change = as.numeric(gsub("[%]", "", market_snapshot_percent_change)),
    hour = lubridate::ymd_hms(hour),
    timestamp = lubridate::ymd_hms(timestamp)
  )
```

### 2.4.2 Handle NA values

For sparse columns like social_manager_alerts, news, agents_unavailable, replace empty lists with NA_character\_.

```{r}
cleaned_df <- cleaned_df %>%
  mutate(across(where(is.character), ~ na_if(.x, ""))) %>%
  select(where(~ !all(is.na(.))))

```

### 2.4.3 Normalize agent roles

Collapse duplicate role columns (agent_role.x, agent_role.y) into one clean agent_role.

```{r}
cleaned_df <- combined_df %>%
  mutate(agent_role = coalesce(agent_role.x, agent_role.y),
         agent_label = coalesce(agent_label.x, agent_label.y)) %>%
  select(-agent_role.x, -agent_role.y, -agent_label.x, -agent_label.y)
```

### 2.4.4 Categorical encoding 

Convert market_snapshot_sentiment into factors for modeling:
```{r}
cleaned_df <- cleaned_df %>%
  mutate(market_snapshot_sentiment = factor(market_snapshot_sentiment,
                                            levels = c("negative","neutral","positive")))
```
### 2.4.5 Derived Features

```{r}
cleaned_df <- cleaned_df %>%
  mutate(
    message_length = str_length(content),
    is_reply = !is.na(responding_to),
    round_day = lubridate::date(hour)
  )
```

### 2.4.6 Text cleaning

```{r}
cleaned_df <- combined_df %>%
  mutate(content = str_squish(content))
```

```{r}
glimpse(cleaned_df)
```

