---
title: "Hands-on_Exercise 1"
author: "Jiang Yuxi"
date: "April 16, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# 1 A Layered Grammar of Graphics: ggplot2 methods

# 1.1 Getting Started

## 1.1.1 Install and launching R packages

The code chunk below uses p_load() of pacman package to check if tidyverse packages are installed in the computer. If they are, then they will be launched into R.

```{r}
pacman::p_load(tidyverse)
```

## 1.1.2 Importing the data

The code chunk below uses readr to import Exam_data.csv.

```{r}
exam_data <- read.csv("data/Exam_data.csv")
```

# 1.2 Introducing ggplot

ggplot is used for data-driven graphics and is a part of the tidyverse family specially designed for visual exploration and communication.

## 1.2.1 R Graphics VS ggplot

The code chunk below uses hist to quick check and generate R Graphics with limited customization, while ggplot needs more setup with declarative style and is best for publication-ready visuals.

R Graphics

```{r}
hist(exam_data$MATHS)
```

ggplot2

```{r}
ggplot(data=exam_data, aes(x = MATHS)) +
  geom_histogram(bins=10, 
                 boundary = 100,
                 color="black", 
                 fill="grey") +
  ggtitle("Distribution of Maths scores")
```

# 1.3 Grammar of Graphics

**Grammar of Graphics** is a general scheme for data visualization which breaks up graphs into semantic components such as scales and layers. It was introduced by Leland Wilkinson (1999) Grammar of Graphics, Springer.

## 1.3.1 A Layered Grammar of Graphics

ggplot2 is an implementation of Leland Wilkinson’s Grammar of Graphics. Figure below shows the **seven grammars** of ggplot2.

A short description of each building block are as follows:

\-**Data**: The dataset being plotted.

\-**Aesthetics** take attributes of the data and use them to influence visual characteristics, such as position, colours, size, shape, or transparency.

\-**Geometrics**: The visual elements used for our data, such as point, bar or line.

\-**Facets** split the data into subsets to create multiple variations of the same graph (paneling, multiple plots).

\-**Statistics**, statiscal transformations that summarise data (e.g. mean, confidence intervals).

\-**Coordinate systems** define the plane on which data are mapped on the graphic.

\-**Themes** modify all non-data components of a plot, such as main title, sub-title, y-aixs title, or legend background.

# 1.4 Essential Grammatical Elements in ggplot2: data

```{r}
ggplot(data=exam_data)
```

# 1.5 Essential Grammatical Elements in ggplot2: Aesthetic mappings

Code chunk below adds the aesthetic element into the plot.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS))
```

# 1.6 Essential Grammatical Elements in ggplot2: geom

Geometric objects are the actual marks we put on a plot. Examples include:

**geom_point** for drawing individual points (e.g., a scatter plot)

**geom_line** for drawing lines (e.g., for a line charts)

**geom_smooth** for drawing smoothed lines (e.g., for simple trends or approximations)

**geom_bar** for drawing bars (e.g., for bar charts)

**geom_histogram** for drawing binned values (e.g. a histogram)

**geom_polygon** for drawing arbitrary shapes geom_map for drawing polygons in the shape of a map! (You can access the data to use for these maps by using the map_data() function).

## 1.6.1 Geometric Objects: geom_bar

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()
```

## 1.6.2 Geometric Objects: geom_dotplot

The y scale is not very useful, in fact it is very misleading.

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(dotsize = 0.5)
```

Adjusted plot

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(binwidth=2.5,         
               dotsize = 0.5) +      
  scale_y_continuous(NULL,           
                     breaks = NULL)  
```

## 1.6.3 Geometric Objects: geom_histogram()

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_histogram()  
```

## 1.6.4 Modifying a geometric object by changing geom()

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20,            
                 color="black",      
                 fill="light blue")  
```

## 1.6.5 Modifying a geometric object by changing aes()

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           fill = GENDER)) +
  geom_histogram(bins=20, 
                 color="grey30")
```

## 1.6.6 Geometric Objects: geom-density()

The code below plots the distribution of Maths scores in a kernel density estimate plot.

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_density()           
```

The code chunk below plots two kernel density lines by using colour or fill arguments of aes()

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS, 
           colour = GENDER)) +
  geom_density()
```

## 1.6.7 Geometric Objects: geom_boxplot

The code chunk below plots boxplots by using geom_boxplot().

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS,       
           x= GENDER)) +    
  geom_boxplot()            
```

The code chunk below plots the distribution of Maths scores by gender in notched plot instead of boxplot.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot(notch=TRUE)
```

## 1.6.8 Geometric Objects: geom_violin

The code below plot the distribution of Maths score by gender in violin plot.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_violin()
```

## 1.6.9 Geometric Objects: geom_point()

The code chunk below plots a scatterplot showing the Maths and English grades of pupils by using geom_point().

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point()            
```

## 1.6.10 geom objects can be combined

The code chunk below plots the data points on the boxplots by using both geom_boxplot() and geom_point().

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot() +                    
  geom_point(position="jitter", 
             size = 0.5)        
```

# 1.7 Essential Grammatical Elements in ggplot2: stat

The Statistics functions statistically transform data, usually as some form of summary. For example:

frequency of values of a variable (bar graph) - a mean - a confidence limit

There are two ways to use these functions: add a stat\_() function and override the default geom, or add a geom\_() function and override the default stat.

## 1.7.1 Working with stat()

The boxplots below are incomplete because the positions of the means were not shown.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot()
```

## 1.7.2 Working with stat - the stat_summary() method

The code chunk below adds mean values by using stat_summary() function and overriding the default geom.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  stat_summary(geom = "point",       
               fun = "mean",         
               colour ="red",        
               size=4)               
```

## 1.7.3 Working with stat - the geom() method

The code chunk below adding mean values by using geom\_() function and overriding the default stat.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  geom_point(stat="summary",        
             fun="mean",           
             colour="red",          
             size=4)          
```

## 1.7.4 Adding a best fit curve on a scatterplot?

The scatterplot below shows the relationship of Maths and English grades of pupils. The interpretability of this graph can be improved by adding a best fit curve.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(size=0.5)
```

The default smoothing method can be overridden as shown below.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              linewidth=0.5)
```

# 1.8 Essential Grammatical Elements in ggplot2: Facets

Facetting generates small multiples (sometimes also called trellis plot), each displaying a different subset of the data. They are an alternative to aesthetics for displaying additional discrete variables. ggplot2 supports two types of factes, namely: facet_grid() and facet_wrap.

## 1.8.1 Working with facet_wrap()

The code chunk below plots a trellis plot using facet-wrap().

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_wrap(~ CLASS)
```

## 1.8.2 facet_grid() function

The code chunk below plots a trellis plot using facet_grid().

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_grid(~ CLASS)
```

# 1.9 Essential Grammatical Elements in ggplot2: Coordinates

## 1.9.1 Working with Coordinate

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()
```

The code chunk below flips the horizontal bar chart into vertical bar chart by using coord_flip().

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip()
```

## 1.9.2 Changing the y- and x-axis range

The scatterplot on the right is slightly misleading because the y-aixs and x-axis range are not equal.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, size=0.5)
```

The code chunk below fixed both the y-axis and x-axis range from 0-100.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
```

# 1.10 Essential Grammatical Elements in ggplot2: themes

Themes control elements of the graph not related to the data. For example:

background colour size of fonts gridlines colour of labels

## 1.10.1 Working with theme

The code chunk below plot a horizontal bar chart using theme_gray().

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_gray()
```

A horizontal bar chart plotted using theme_classic().

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_classic()
```

A horizontal bar chart plotted using theme_minimal().

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_minimal()
```

# Pre-set up example

## Plotting a simple bar chart

```{r}
library(ggplot2)
ggplot(data = exam_data,
       aes(x = RACE)) +
  geom_bar( fill = "forestgreen") +
  labs(
    x = "Race Category",
    y = "Number of Students",
    title = "Distribution of Students by Race"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"))
```
