---
title: "Take-home_Exercise 1 Excecutive Summary"
format: revealjs
---

## Exploring Profitability and Volatility in Motor Insurance

<br><br><br><br><br>

```         
                               Jiang Yuxi 
                             April Term 2026 
```

## Contents

::: {style="font-size:0.8em; line-height:1.1"}
1.  Cost Distribution

2.  Premium Distribution

3.  Claim Distribution

4.  Profitability Metrics

5.  Factor Distribution

6.  Correlation Analysis

7.  Profitability vs Risk Metrics

8.  Policy & Vehicle Characteristics

9.  Driver Age vs Claim Severity

10. Executive Summary
:::

```{r}
pacman::p_load(tidyverse, readr, dplyr, patchwork, zoo, ggridges, plotly,
               forcats, skimr, scales, corrplot, ggiraph) 
```

```{r}
portfolio <- read_delim("data/Dataset of motor insurance portfolio.csv", delim = ";")
```

```{r}
portfolio <- portfolio %>%
  mutate(
    # columns → integer
    insured_id = as.integer(insured_id),
    year = as.integer(year),
    driver_age = as.integer(driver_age),
    vehicle_age = as.integer(vehicle_age),
    age_driving_licence = as.integer(age_driving_licence),
    seats = as.integer(seats),
    total_claims = as.integer(total_claims),
    liability_claims = as.integer(liability_claims),
    liability_property_claims = as.integer(liability_property_claims),
    liability_injury_claims = as.integer(liability_injury_claims),
    property_claims = as.integer(property_claims),
    theft_claims = as.integer(theft_claims),
    fire_claims = as.integer(fire_claims),
    glass_claims = as.integer(glass_claims),
    legal_protection_claims = as.integer(legal_protection_claims),
    occupants_claims = as.integer(occupants_claims),
    
    # columns → factor
    policy_type = factor(policy_type),
    policy_status = factor(policy_status),
    business_type = factor(business_type),
    payment_frequency = factor(payment_frequency),
    bonus_score = factor(bonus_score),
    vehicle_brand = factor(vehicle_brand),
    fuel_type = factor(fuel_type),
    municipality_type = factor(municipality_type),
    circulation_area = factor(circulation_area)
  )
```

```{r}
# Precompute threshold for high-cost claims
threshold_incurred <- quantile(portfolio$total_incurred, 0.95, na.rm = TRUE)

portfolio <- portfolio %>%
  mutate(
    # Profitability measures
    loss_ratio = ifelse(total_premium > 0, total_incurred / total_premium, NA_real_),
    profit = total_premium - total_incurred,
    
    # Risk metrics
    claim_frequency = ifelse(total_exposure > 0, total_claims / total_exposure, 0),  # set 0 if exposure=0
    claim_severity = ifelse(total_claims > 0, total_incurred / total_claims, NA_real_),
    
    # Exposure-adjusted metrics
    premium_per_exposure = ifelse(total_exposure > 0, total_premium / total_exposure, NA_real_),
    incurred_per_exposure = ifelse(total_exposure > 0, total_incurred / total_exposure, NA_real_),
    
    # Binary indicators as factors
    has_claim = factor(ifelse(total_claims > 0, 1, 0), levels = c(0,1), labels = c("No Claim","Has Claim")),
    high_cost_claim = factor(ifelse(total_incurred > threshold_incurred, 1, 0), levels = c(0,1), labels = c("Normal","High Cost")),
    profitable_policy = factor(ifelse(profit > 0, 1, 0), levels = c(0,1), labels = c("Not Profitable","Profitable"))
  )
```

```{r}
portfolio <- portfolio %>%
  mutate(
    # Impute vehicle_value with brand-specific median
    vehicle_value = ifelse(is.na(vehicle_value),
                           ave(vehicle_value, vehicle_brand,
                               FUN = function(x) median(x, na.rm = TRUE)),
                           vehicle_value),
    
    # Impute fuel_type with "Unknown"
    fuel_type = fct_explicit_na(factor(fuel_type), na_level = "Unknown")
  )
    
# Drop rows with missing driver/vehicle age
portfolio <- portfolio %>%
  filter(!is.na(vehicle_age), !is.na(age_driving_licence))
```

```{r}
portfolio <- portfolio %>%
  mutate(
    # Flags for structural NA cases
    has_premium = total_premium > 0,
    has_exposure = total_exposure > 0,
    has_claims = total_claims > 0
  )
```

```{r}
# Step 1: Compute IQR thresholds for loss_ratio
Q1_lr <- quantile(portfolio$loss_ratio, 0.25, na.rm = TRUE)
Q3_lr <- quantile(portfolio$loss_ratio, 0.75, na.rm = TRUE)
IQR_lr <- Q3_lr - Q1_lr

# Step 2: Compute IQR thresholds for profit
Q1_pf <- quantile(portfolio$profit, 0.25, na.rm = TRUE)
Q3_pf <- quantile(portfolio$profit, 0.75, na.rm = TRUE)
IQR_pf <- Q3_pf - Q1_pf

# Step 3: Flag outliers for these two columns
portfolio <- portfolio %>%
  mutate(
    is_outlier_loss_ratio = loss_ratio < (Q1_lr - 1.5 * IQR_lr) | loss_ratio > (Q3_lr + 1.5 * IQR_lr),
    is_outlier_profit     = profit < (Q1_pf - 1.5 * IQR_pf) | profit > (Q3_pf + 1.5 * IQR_pf),
    is_outlier            = is_outlier_loss_ratio | is_outlier_profit
  )

# Step 4: Drop rows flagged as outliers
portfolio_drop <- portfolio %>%
  filter(!is_outlier)
```

```{r}
portfolio <- portfolio %>%
  mutate(
    policy_status = recode(policy_status,
      "A" = "Active",
      "C" = "Cancelled"
    ),

    business_type = recode(business_type,
      "NB" = "New Business",
      "P" = "Portfolio"
    ),

    payment_frequency = recode(payment_frequency,
      "S" = "Semi-annual",
      "Q" = "Quarterly",
      "A" = "Annual"
    ),

    bonus_score = recode(bonus_score,
      "G" = "Good",
      "N" = "Neutral",
      "B" = "Bad"
    ),

    fuel_type = recode(fuel_type,
      "D" = "Diesel",
      "G" = "Gasoline"
    ),

    municipality_type = recode(municipality_type,
      "IS" = "Islands",
      "I" = "Inland",
      "C" = "Coastal"
    ),

    circulation_area = recode(circulation_area,
      "U" = "Urban",
      "R" = "Rural"
    )
  )
```

## 1. Cost Distribution

```{r}
#| echo: false
#｜ fig-width: 6
#｜ fig_height: 4
#| out.width: "60%"
# Reshape incurred cost data into long format
incurred_clean <- portfolio %>%
  select(
    liability_incurred,
    liability_property_incurred,
    liability_injury_incurred,
    property_incurred,
    theft_incurred,
    fire_incurred,
    glass_incurred,
    legal_protection_incurred,
    occupants_incurred
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "claim_type",
    values_to = "cost"
  ) %>%
  filter(is.finite(cost), cost > 0)

# ECDF transformation per claim type
cdf_data <- incurred_clean %>%
  group_by(claim_type) %>%
  arrange(cost) %>%
  mutate(
    cum_pct = row_number() / n() * 100
  ) %>%
  ungroup()

# Plot cumulative distribution
ggplot(cdf_data, aes(x = cum_pct, y = cost, color = claim_type)) +
  geom_line(linewidth = 1, alpha = 0.8) +
  scale_y_log10(
    labels = scales::comma,
    limits = c(1.0, NA) ) +
  labs(
    title = "Cumulative Distribution of Incurred Costs by Claim Type",
    x = "Cumulative Percentage of Claims (%)",
    y = "Incurred Cost (log scale)",
    color = "Claim Type"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 8),
    plot.title = element_text(face = "bold"),
    plot.margin = margin(10, 10, 30, 10)
  ) +
  guides(color = guide_legend(nrow = 3, ncol = 3, byrow = TRUE))
```

::: {style="font-size:0.5em; margin-top:10px"}
**Insight:** The plot reveals that certain claim types, such as liability injury and fire claims, have much higher incurred costs at the upper end of the distribution compared to others. This indicates that while most claims are relatively small, a small proportion of catastrophic claims drive portfolio volatility and can heavily impact profitability.
:::

::: notes
Here we see the distribution of incurred costs across claim types. Most claims are relatively small, but liability injury and fire claims stand out with much higher costs at the upper end. This tells us that while everyday claims are manageable, catastrophic ones are the real drivers of volatility and profitability risk.
:::

## 2. Premium Distribution

```{r}
#| echo: false
#｜ fig-width: 6
#｜ fig_height: 4
#| out.width: "60%"
# Reshape incurred cost data into long format
incurred_long <- portfolio %>%
  select(
    liability_premium,
    property_damage_premium,
    theft_premium,
    fire_premium,
    glass_premium,
    legal_protection_premium,
    occupants_premium
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "insurance_type",
    values_to = "premium"
  )

# Stacked density plot
ggplot(incurred_long %>% filter(!is.na(premium), premium > 0),
       aes(x = premium, fill = insurance_type)) +
  geom_density(alpha = 0.6, position = "stack") +
  scale_x_log10(labels = comma) +   # log scale to handle skewness
  theme_minimal() +
  labs(
    title = "Stacked Density Plot of Premium",
    x = "Premuim (log scale)",
    y = "Density"
  ) +
  theme(
    strip.text = element_text(face = "bold"),
    legend.title = element_blank()
  )
```

::: {style="font-size:0.5em; margin-top:10px"}
**Insight:** The plot shows that most premiums cluster at lower values, but the log scale reveals long tails where certain insurance types extend into higher premium ranges. Liability and property damage premiums dominate the distribution, while niche coverages like glass or legal protection appear less frequent and concentrated. This highlights that core coverages drive the bulk of premium income, whereas specialized coverages contribute smaller but distinct segments.
:::

::: notes
Premiums are mostly concentrated at lower values. However, liability and property damage premiums dominate the portfolio, while niche coverages like glass or legal protection are smaller but distinct. This highlights that core coverages generate the bulk of income, while specialized ones add diversity.
:::

## 3. Claim Distribution

```{r}
#| echo: false
#｜ fig-width: 6
#｜ fig_height: 4
#| out.width: "60%"
# Reshape claim count columns into long format
claims_long <- portfolio %>%
  select(
    liability_claims,
    liability_property_claims,
    liability_injury_claims,
    property_claims,
    theft_claims,
    fire_claims,
    glass_claims,
    legal_protection_claims,
    occupants_claims
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "claim_type",
    values_to = "claim_count"
  ) %>%
  filter(is.finite(claim_count))

# Plot boxplots
ggplot(claims_long, aes(x = claim_type, y = claim_count, fill = claim_type)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.4) +
  labs(
    title = "Distribution of Claim Counts by Type",
    x = "Claim Type",
    y = "Number of Claims"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold")
  )
```

::: {style="font-size:0.5em; margin-top:10px"}
**Insight:** The plot shows that most claim types have relatively low counts, with the majority clustering near zero, but property claims and liability‑related claims tend to have higher variability and more frequent occurrences. This indicates that property and liability claims are a major driver of claim activity, while categories like fire or glass are less common but still contribute to overall portfolio risk.
:::

::: notes
Looking at claim counts, most categories cluster near zero. But property and liability claims show higher variability and frequency. These are the main drivers of claim activity, while categories like fire or glass are less common but still contribute to overall risk.
:::

## 4. Profitability Metrics

```{r}
#| echo: false
#｜ fig-width: 6
#｜ fig_height: 4
#| out.width: "60%"
portfolio_bin <- portfolio %>%
  filter(has_premium) %>%
  mutate(
    # Loss ratio bins
    loss_ratio_bin = case_when(
      loss_ratio == 0 ~ "=0",
      loss_ratio > 0 & loss_ratio <= 1 ~ "0-1",
      loss_ratio > 1 ~ ">1",
      TRUE ~ NA_character_
    ),
    # Profit bins
    profit_bin = case_when(
      profit < 0 ~ "Negative",
      profit >= 0 & profit < 250 ~ "0-250",
      profit >= 250 & profit < 500 ~ "250-500",
      profit >= 500 ~ ">500",
      TRUE ~ NA_character_
    )
  )

p1 <- ggplot(portfolio_bin, aes(x = loss_ratio_bin, fill = loss_ratio_bin)) +
  geom_bar(alpha = 0.7) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  labs(title = "Loss Ratio Bins", x = "Loss Ratio Category", y = "Count") +
  scale_x_discrete(drop = TRUE) +
  theme_minimal() +
  theme(legend.position = "none")

p2 <- ggplot(portfolio_bin, aes(x = profit_bin, fill = profit_bin)) +
  geom_bar(alpha = 0.7) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  labs(title = "Profit Bins", x = "Profit Category", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")

p1 + p2
```

::: {style="font-size:0.5em; margin-top:10px"}
**Insight:** The charts show that a large majority of policies have a loss ratio of zero, meaning no claims were made, while a smaller but significant group has loss ratios greater than one, indicating unprofitable policies where claims exceeded premiums. On the profit side, most policies fall into moderate profit ranges (0–250 and 250–500), but there is also a notable share of highly profitable policies (\>500). This highlights that while many policies generate steady profits, a subset of high‑loss policies erodes overall portfolio performance.
:::

::: notes
On the profitability side, many policies have no claims at all, but some exceed premiums, making them unprofitable. Most policies generate moderate profits, yet a subset of high-loss policies erodes overall performance. This mix of steady profits and risky outliers is a key challenge.
:::

## 5. Categorical Variables Distribution

```{r}
#| echo: false
#｜ fig-width: 6
#｜ fig_height: 4
#| out.width: "60%"
# Categorical variables: bar charts
# Select categorical variables
cat_long <- portfolio %>%
  select(
    policy_type,
    policy_status,
    business_type,
    payment_frequency,
    bonus_score,
    fuel_type,
    municipality_type,
    circulation_area
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  )

# Create percentage table within each variable
plot_data <- cat_long %>%
  group_by(variable, value) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(variable) %>%
  mutate(
    percent = count / sum(count),
    value = fct_reorder(value, percent)
  )

# Plot
ggplot(plot_data,
       aes(x = percent,
           y = value)) +
  geom_col(fill = "steelblue") +
  facet_wrap(~variable,
             scales = "free_y") +
  scale_x_continuous(labels = percent_format()) +
  labs(
    title = "Distribution of Categorical Variables",
    x = "Percentage",
    y = NULL
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold"),
    axis.text.y = element_text(size = 9)
  )
```

::: {style="font-size:0.5em; margin-top:10px"}
**Insight:** The charts show that most policies are active and concentrated in portfolio business rather than new business, with annual payment frequency being the most common. Bonus scores lean heavily toward “Good,” while fuel types are dominated by gasoline, though a notable share is marked “Unknown.” Urban circulation areas and coastal municipalities are more prevalent, reflecting demographic and geographic concentration. These distributions highlight the portfolio’s structural composition and help explain underlying profitability and risk patterns.
:::

::: notes
Structurally, most policies are active, annual payments are common, and bonus scores lean toward ‘Good.’ Gasoline dominates fuel type, though some are marked ‘Unknown.’ Urban and coastal areas are more prevalent. These distributions give us a snapshot of the portfolio’s composition.
:::

## 6. Correlation Analysis

:::: {.columns}

::: {.column width="50%"}
```{r}
#| echo: false
#｜ fig-width: 8
#｜ fig_height: 6
#| out.width: "100%"
# Select numeric variables
num_vars <- portfolio %>%
  select(total_premium, total_incurred, profit, loss_ratio,
         claim_frequency, claim_severity, total_claims,
         premium_per_exposure, incurred_per_exposure,
         driver_age, vehicle_age, power_to_weight_ratio, vehicle_value)

# Correlation matrix
cor_matrix <- cor(num_vars, use = "complete.obs", method = "spearman")

# Visualize
corrplot.mixed(cor_matrix,
               lower = "ellipse",
               upper = "number",
               number.cex = 0.5,
               tl.cex = 0.8,
               order = "hclust",
               tl.pos = "lt",
               diag = "l",
               tl.col = "black",
               na.label = " ")   
```
:::

::: {.column width="50%"}
```{r}
#| echo: false
cor_df <- cor_matrix %>%
  as.data.frame() %>%
  rownames_to_column("var1") %>%
  pivot_longer(-var1, names_to = "var2", values_to = "correlation") %>%
  filter(var1 < var2, abs(correlation) >= 0.5) %>%
  arrange(desc(abs(correlation)))

ggplot(cor_df, aes(x = reorder(paste(var1, var2, sep = "–"), abs(correlation)),
                   y = correlation, fill = correlation > 0)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(title = "Top Correlations (|ρ| ≥ 0.5)",
       x = "Variable Pair",
       y = "Spearman Correlation") +
  theme_minimal()
```
:::
::::

::: {style="font-size:0.5em; margin-top:10px"}
**Insight:** Profitability is most negatively affected by high incurred costs and loss ratios (ρ ≈ –0.93 to –0.94), showing that claims relative to premiums directly erode profits. Claim severity is strongly correlated with incurred costs and loss ratio (ρ ≈ 0.87–0.91), highlighting that catastrophic claims destabilize the portfolio more than claim frequency. Claim frequency is highly correlated with loss ratio (ρ ≈ 0.90, p \< 2.2e‑16), confirming that frequent claims drive unprofitability. Demographic factors like driver age and vehicle age show moderate correlation (ρ ≈ 0.86), but financial ratios remain the dominant drivers of portfolio performance.
:::

::: notes
Correlation analysis shows that profitability is most negatively affected by incurred costs and loss ratios. Claim severity destabilizes the portfolio more than frequency, but frequent claims still drive unprofitability. Demographics like driver and vehicle age matter moderately, yet financial ratios remain the dominant drivers.
:::

## 7. Profitability vs Risk Metrics

```{r}
#| echo: false
#｜ fig-width: 6
#｜ fig_height: 4
#| out.width: "60%"
# Premium vs Incurred (profitability)

portfolio_clean <- portfolio %>%
  filter(total_incurred > 0, total_premium >= 0)

p1 <- ggplot(portfolio, aes(x = total_premium, y = total_incurred)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  scale_y_log10() +
  coord_cartesian(xlim = c(0, 3000), ylim = c(1, 500000)) +  
  geom_smooth(method = "lm", se = TRUE, color = "darkred") +
  labs(
    title = "Premium vs Incurred Costs",
    x = "Total Premium",
    y = "Total Incurred"
  ) +
  theme_minimal(base_size = 14)

p2 <- ggplot(portfolio %>% filter(has_premium), aes(x = claim_frequency, y = loss_ratio)) +
  geom_point(alpha = 0.4, color = "steelblue") +   # scatter points
  coord_cartesian(xlim = c(0, 100), ylim = c(0, 1000)) +
  geom_smooth(method = "lm", se = TRUE, color = "darkred") +  # regression line
  labs(
    title = "Loss Ratio vs Claim Frequency",
    x = "Claim Frequency (claims per exposure)",
    y = "Loss Ratio"
  ) +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"))

p1+p2
```

::::::: columns
:::: {.column width="50%"}
::: {style="font-size:0.5em; line-height:1.0"}
**Insight (Premiums vs Costs)**\
Higher premiums are linked to higher incurred costs. Most policies stay low, but a few high‑premium ones drive large claims and greater volatility.
:::
::::

:::: {.column width="50%"}
::: {style="font-size:0.5em; line-height:1.0"}
**Insight (Claims vs Loss)**\
Frequent claims push loss ratios up. Even small claims erode profitability when they occur often, making claim frequency a key driver of portfolio stability.
:::
::::
:::::::

::: notes
Here we compare premiums with incurred costs and claim frequency with loss ratios. Higher premiums are linked to higher incurred costs, with volatility driven by large claims. Frequent claims push loss ratios up, even when individual claims are small. Both relationships confirm the key drivers of portfolio risk.
:::

## 8. Policy & Vehicle Characteristics

```{r}
#| echo: false
#｜ fig-width: 6
#｜ fig_height: 4
#| out.width: "60%"
# Mean profit per policy type with confidence intervals
ggplot(portfolio, aes(
    x = fct_reorder(policy_type, profit, .fun = mean, .desc = TRUE),
    y = profit,
    fill = policy_type
  )) +
  stat_summary(fun = mean, geom = "bar", alpha = 0.7) +
  stat_summary(fun.data = mean_cl_boot, geom = "errorbar", width = 0.2) +
  stat_summary(
    fun = mean,
    geom = "text",
    aes(label = scales::comma(..y..)),
    vjust = -0.5, size = 3.5, color = "black"
  ) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Mean Profit by Policy Type",
    x = "Policy Type",
    y = "Mean Profit"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold")
  )
```

::: {style="font-size:0.5em; margin-top:10px"}
**Insight:**

\- COMP_E policies generate the highest mean profit, suggesting they are the most financially favorable product line.

\- COMP_N policies show the lowest mean profit, indicating weaker performance and potential risk to overall profitability.

\- Other policy types (CC, TP, TPG) fall in between, with moderate profitability compared to COMP_E.

\- The confidence intervals highlight variability, but the overall ranking of policy types remains consistent, guiding strategic focus toward more profitable products.
:::

::: notes
Policy type matters. COMP_E policies generate the highest mean profit, making them the most favorable. COMP_N policies show the weakest performance, while others fall in between. Despite variability, the ranking is consistent, guiding us toward focusing on profitable product lines.
:::

## 9. Driver Age vs Claim Severity

```{r}
#| echo: false
#｜ fig-width: 6
#｜ fig_height: 4
#| out.width: "60%"
portfolio_age <- portfolio %>% 
  mutate(
    age_band = cut(driver_age,
                   breaks = seq(18, 80, by = 10),
                   include.lowest = TRUE)
  ) %>%
  filter(!is.na(age_band)) %>%   # <-- remove NA age bands
  group_by(age_band) %>%
  mutate(
    p25   = quantile(claim_severity, 0.25, na.rm = TRUE),
    q3    = quantile(claim_severity, 0.75, na.rm = TRUE),
    iqr   = q3 - p25,
    lower = p25,                  
    upper = q3 + 1.5 * iqr        
  ) %>%
  ungroup() %>%
  filter(claim_severity >= lower, claim_severity <= upper)

# Ridge plot of loss ratio by age band (log scale, NA removed)
ggplot(portfolio_age, aes(x = loss_ratio, y = age_band, fill = age_band)) +
  ggridges::geom_density_ridges(alpha = 0.6) +
  scale_x_log10() +
  labs(
    title = "Claim Severity Distribution by Driver Age Band (P25 cutoff, log scale)",
    x = "Log Claim Severity",
    y = "Driver Age Band"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```

::: {style="font-size:0.5em; margin-top:10px"}
**Insight:** Claim severity is most stable in mid‑age groups, while both younger and older drivers add more volatility.

\- Younger drivers (18–28) show slightly higher severity risk

\- Middle‑aged drivers (38–58) are more stable and predictable.

\- Older drivers (68–78) again show wider variability.
:::

::: notes
Claim severity varies by driver age. Middle-aged drivers are the most stable, while younger and older drivers show more volatility. This pattern suggests that age bands influence risk, with mid-age groups offering more predictable outcomes.
:::

## 10. Executive Summary

::: {style="font-size:0.5em; line-height:1.1"}
**Key Findings:**

**Claims behavior drives volatility.** Claim frequency is highly unstable relative to its mean (CV ≈ 6.17), while claim severity also contributes (CV ≈ 3.45). Catastrophic claims amplify incurred costs, while frequent smaller claims steadily erode profitability.

**Policy type and bonus score are critical differentiators.** COMP_E policies show the highest loss ratio variance, indicating unstable underwriting results, while COMP_N policies are more predictable. Bonus scores significantly influence profitability outcomes.

**Geography and fuel type play a limited role.** Statistical tests show no significant differences in variance or mean loss ratios across circulation areas or fuel types, suggesting these variables are less impactful in explaining volatility.

**Portfolio risk is dynamic over time.** Rolling time‑series volatility reveals spikes linked to shocks or seasonal effects, confirming that risk is not constant and requires ongoing monitoring.

**Visual analytics adds measurable value.** Univariate plots highlight skewness and outliers, bivariate analyses reveal correlations between premiums, incurred costs, and claims, while statistical tests validate which categorical differences are meaningful.

**Conclusion:**\
The investigation confirms that profitability and volatility in motor insurance portfolios are primarily shaped by claims frequency and severity, with policy type and bonus score acting as key structural differentiators. Geography and fuel type are less influential, while portfolio risk fluctuates over time. By combining visual analytics with statistical validation, insurers can move beyond descriptive reporting to actionable insights—targeting unstable product types, managing claim drivers, leveraging bonus scores for segmentation, and monitoring volatility trends to anticipate shocks.
:::

::: notes
To conclude: claims frequency and severity are the main drivers of volatility. Policy type and bonus score are critical differentiators, while geography and fuel type play limited roles. Risk is dynamic over time, requiring ongoing monitoring. Visual analytics adds measurable value by turning descriptive reporting into actionable insights. The strategic takeaway is clear: manage claim drivers, target unstable product types, leverage bonus scores, and monitor volatility trends to anticipate shocks.
:::
