---
title: "Take-home_Exercise 1 Excecutive Summary"
format: revealjs
---

## Take-home Exercise 1

### Exploring Profitability and Volatility in Motor Insurance

<br><br><br><br><br>

```         
                    Jiang Yuxi April Term 2026 
```

## Contents

Insight 1:

Insight 2:

Insight 3:

Insight 4:

Insight 5:

Executive Summary

```{r}
pacman::p_load(tidyverse, readr, dplyr, patchwork, zoo, ggridges, plotly,
               forcats, skimr, scales, corrplot, ggiraph) 
```

```{r}
portfolio <- read_delim("data/Dataset of motor insurance portfolio.csv", delim = ";")
```

```{r}
portfolio <- portfolio %>%
  mutate(
    # columns → integer
    insured_id = as.integer(insured_id),
    year = as.integer(year),
    driver_age = as.integer(driver_age),
    vehicle_age = as.integer(vehicle_age),
    age_driving_licence = as.integer(age_driving_licence),
    seats = as.integer(seats),
    total_claims = as.integer(total_claims),
    liability_claims = as.integer(liability_claims),
    liability_property_claims = as.integer(liability_property_claims),
    liability_injury_claims = as.integer(liability_injury_claims),
    property_claims = as.integer(property_claims),
    theft_claims = as.integer(theft_claims),
    fire_claims = as.integer(fire_claims),
    glass_claims = as.integer(glass_claims),
    legal_protection_claims = as.integer(legal_protection_claims),
    occupants_claims = as.integer(occupants_claims),
    
    # columns → factor
    policy_type = factor(policy_type),
    policy_status = factor(policy_status),
    business_type = factor(business_type),
    payment_frequency = factor(payment_frequency),
    bonus_score = factor(bonus_score),
    vehicle_brand = factor(vehicle_brand),
    fuel_type = factor(fuel_type),
    municipality_type = factor(municipality_type),
    circulation_area = factor(circulation_area)
  )
```

```{r}
# Precompute threshold for high-cost claims
threshold_incurred <- quantile(portfolio$total_incurred, 0.95, na.rm = TRUE)

portfolio <- portfolio %>%
  mutate(
    # Profitability measures
    loss_ratio = ifelse(total_premium > 0, total_incurred / total_premium, NA_real_),
    profit = total_premium - total_incurred,
    
    # Risk metrics
    claim_frequency = ifelse(total_exposure > 0, total_claims / total_exposure, 0),  # set 0 if exposure=0
    claim_severity = ifelse(total_claims > 0, total_incurred / total_claims, NA_real_),
    
    # Exposure-adjusted metrics
    premium_per_exposure = ifelse(total_exposure > 0, total_premium / total_exposure, NA_real_),
    incurred_per_exposure = ifelse(total_exposure > 0, total_incurred / total_exposure, NA_real_),
    
    # Binary indicators as factors
    has_claim = factor(ifelse(total_claims > 0, 1, 0), levels = c(0,1), labels = c("No Claim","Has Claim")),
    high_cost_claim = factor(ifelse(total_incurred > threshold_incurred, 1, 0), levels = c(0,1), labels = c("Normal","High Cost")),
    profitable_policy = factor(ifelse(profit > 0, 1, 0), levels = c(0,1), labels = c("Not Profitable","Profitable"))
  )
```

```{r}
portfolio <- portfolio %>%
  mutate(
    # Impute vehicle_value with brand-specific median
    vehicle_value = ifelse(is.na(vehicle_value),
                           ave(vehicle_value, vehicle_brand,
                               FUN = function(x) median(x, na.rm = TRUE)),
                           vehicle_value),
    
    # Impute fuel_type with "Unknown"
    fuel_type = fct_explicit_na(factor(fuel_type), na_level = "Unknown")
  )
    
# Drop rows with missing driver/vehicle age
portfolio <- portfolio %>%
  filter(!is.na(vehicle_age), !is.na(age_driving_licence))
```

```{r}
portfolio <- portfolio %>%
  mutate(
    # Flags for structural NA cases
    has_premium = total_premium > 0,
    has_exposure = total_exposure > 0,
    has_claims = total_claims > 0
  )
```

```{r}
# Step 1: Compute IQR thresholds for loss_ratio
Q1_lr <- quantile(portfolio$loss_ratio, 0.25, na.rm = TRUE)
Q3_lr <- quantile(portfolio$loss_ratio, 0.75, na.rm = TRUE)
IQR_lr <- Q3_lr - Q1_lr

# Step 2: Compute IQR thresholds for profit
Q1_pf <- quantile(portfolio$profit, 0.25, na.rm = TRUE)
Q3_pf <- quantile(portfolio$profit, 0.75, na.rm = TRUE)
IQR_pf <- Q3_pf - Q1_pf

# Step 3: Flag outliers for these two columns
portfolio <- portfolio %>%
  mutate(
    is_outlier_loss_ratio = loss_ratio < (Q1_lr - 1.5 * IQR_lr) | loss_ratio > (Q3_lr + 1.5 * IQR_lr),
    is_outlier_profit     = profit < (Q1_pf - 1.5 * IQR_pf) | profit > (Q3_pf + 1.5 * IQR_pf),
    is_outlier            = is_outlier_loss_ratio | is_outlier_profit
  )

# Step 4: Drop rows flagged as outliers
portfolio_drop <- portfolio %>%
  filter(!is_outlier)
```

```{r}
portfolio <- portfolio %>%
  mutate(
    policy_status = recode(policy_status,
      "A" = "Active",
      "C" = "Cancelled"
    ),

    business_type = recode(business_type,
      "NB" = "New Business",
      "P" = "Portfolio"
    ),

    payment_frequency = recode(payment_frequency,
      "S" = "Semi-annual",
      "Q" = "Quarterly",
      "A" = "Annual"
    ),

    bonus_score = recode(bonus_score,
      "G" = "Good",
      "N" = "Neutral",
      "B" = "Bad"
    ),

    fuel_type = recode(fuel_type,
      "D" = "Diesel",
      "G" = "Gasoline"
    ),

    municipality_type = recode(municipality_type,
      "IS" = "Islands",
      "I" = "Inland",
      "C" = "Coastal"
    ),

    circulation_area = recode(circulation_area,
      "U" = "Urban",
      "R" = "Rural"
    )
  )
```

## Insight 1: Cost Distribution

```{r}
#| echo: false
#｜ fig-width: 6
#｜ fig_height: 4
#| out.width: "60%"
# Reshape incurred cost data into long format
incurred_clean <- portfolio %>%
  select(
    liability_incurred,
    liability_property_incurred,
    liability_injury_incurred,
    property_incurred,
    theft_incurred,
    fire_incurred,
    glass_incurred,
    legal_protection_incurred,
    occupants_incurred
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "claim_type",
    values_to = "cost"
  ) %>%
  filter(is.finite(cost), cost > 0)

# ECDF transformation per claim type
cdf_data <- incurred_clean %>%
  group_by(claim_type) %>%
  arrange(cost) %>%
  mutate(
    cum_pct = row_number() / n() * 100
  ) %>%
  ungroup()

# Plot cumulative distribution
ggplot(cdf_data, aes(x = cum_pct, y = cost, color = claim_type)) +
  geom_line(linewidth = 1, alpha = 0.8) +
  scale_y_log10(
    labels = scales::comma,
    limits = c(1.0, NA) ) +
  labs(
    title = "Cumulative Distribution of Incurred Costs by Claim Type",
    x = "Cumulative Percentage of Claims (%)",
    y = "Incurred Cost (log scale)",
    color = "Claim Type"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 8),
    plot.title = element_text(face = "bold"),
    plot.margin = margin(10, 10, 30, 10)
  ) +
  guides(color = guide_legend(nrow = 3, ncol = 3, byrow = TRUE))
```

::: {style="font-size:0.9em; margin-top:10px"}
**Insight:** Heavier cars tend to have lower fuel efficiency, as shown by the negative correlation between weight and mpg.
:::

## Insight 2: Premium Distribution

```{r}
#| echo: false
#｜ fig-width: 6
#｜ fig_height: 4
#| out.width: "60%"
# Reshape incurred cost data into long format
incurred_long <- portfolio %>%
  select(
    liability_premium,
    property_damage_premium,
    theft_premium,
    fire_premium,
    glass_premium,
    legal_protection_premium,
    occupants_premium
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "insurance_type",
    values_to = "premium"
  )

# Stacked density plot
ggplot(incurred_long %>% filter(!is.na(premium), premium > 0),
       aes(x = premium, fill = insurance_type)) +
  geom_density(alpha = 0.6, position = "stack") +
  scale_x_log10(labels = comma) +   # log scale to handle skewness
  theme_minimal() +
  labs(
    title = "Stacked Density Plot of Premium",
    x = "Premuim (log scale)",
    y = "Density"
  ) +
  theme(
    strip.text = element_text(face = "bold"),
    legend.title = element_blank()
  )
```

## Insight 3: Claim Distribution

```{r}
#| echo: false
#｜ fig-width: 6
#｜ fig_height: 4
#| out.width: "60%"
# Reshape claim count columns into long format
claims_long <- portfolio %>%
  select(
    liability_claims,
    liability_property_claims,
    liability_injury_claims,
    property_claims,
    theft_claims,
    fire_claims,
    glass_claims,
    legal_protection_claims,
    occupants_claims
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "claim_type",
    values_to = "claim_count"
  ) %>%
  filter(is.finite(claim_count))

# Plot boxplots
ggplot(claims_long, aes(x = claim_type, y = claim_count, fill = claim_type)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.4) +
  labs(
    title = "Distribution of Claim Counts by Type",
    x = "Claim Type",
    y = "Number of Claims"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold")
  )
```

## Insight 4: Profitability Metrics

```{r}
#| echo: false
#｜ fig-width: 6
#｜ fig_height: 4
#| out.width: "60%"
portfolio_bin <- portfolio %>%
  filter(has_premium) %>%
  mutate(
    # Loss ratio bins
    loss_ratio_bin = case_when(
      loss_ratio == 0 ~ "=0",
      loss_ratio > 0 & loss_ratio <= 1 ~ "0-1",
      loss_ratio > 1 ~ ">1",
      TRUE ~ NA_character_
    ),
    # Profit bins
    profit_bin = case_when(
      profit < 0 ~ "Negative",
      profit >= 0 & profit < 250 ~ "0-250",
      profit >= 250 & profit < 500 ~ "250-500",
      profit >= 500 ~ ">500",
      TRUE ~ NA_character_
    )
  )

p1 <- ggplot(portfolio_bin, aes(x = loss_ratio_bin, fill = loss_ratio_bin)) +
  geom_bar(alpha = 0.7) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  labs(title = "Loss Ratio Bins", x = "Loss Ratio Category", y = "Count") +
  scale_x_discrete(drop = TRUE) +
  theme_minimal() +
  theme(legend.position = "none")

p2 <- ggplot(portfolio_bin, aes(x = profit_bin, fill = profit_bin)) +
  geom_bar(alpha = 0.7) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  labs(title = "Profit Bins", x = "Profit Category", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")

p1 + p2
```

## Insight 5: Categorical Variables Distribution

```{r}
#| echo: false
#｜ fig-width: 6
#｜ fig_height: 4
#| out.width: "60%"
# Categorical variables: bar charts
# Select categorical variables
cat_long <- portfolio %>%
  select(
    policy_type,
    policy_status,
    business_type,
    payment_frequency,
    bonus_score,
    fuel_type,
    municipality_type,
    circulation_area
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  )

# Create percentage table within each variable
plot_data <- cat_long %>%
  group_by(variable, value) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(variable) %>%
  mutate(
    percent = count / sum(count),
    value = fct_reorder(value, percent)
  )

# Plot
ggplot(plot_data,
       aes(x = percent,
           y = value)) +
  geom_col(fill = "steelblue") +
  facet_wrap(~variable,
             scales = "free_y") +
  scale_x_continuous(labels = percent_format()) +
  labs(
    title = "Distribution of Categorical Variables",
    x = "Percentage",
    y = NULL
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold"),
    axis.text.y = element_text(size = 9)
  )
```

```         
::: Speaker Notes

:::
```

## Executive Summary

::: {style="font-size:0.8em; line-height:1.3"}
**Key Findings:**

**Claims behavior drives volatility.** Claim frequency is highly unstable relative to its mean (CV ≈ 6.17), while claim severity also contributes (CV ≈ 3.45). Catastrophic claims amplify incurred costs, while frequent smaller claims steadily erode profitability.

**Policy type and bonus score are critical differentiators.** COMP_E policies show the highest loss ratio variance, indicating unstable underwriting results, while COMP_N policies are more predictable. Bonus scores significantly influence profitability outcomes.

**Geography and fuel type play a limited role.** Statistical tests show no significant differences in variance or mean loss ratios across circulation areas or fuel types, suggesting these variables are less impactful in explaining volatility.

**Portfolio risk is dynamic over time.** Rolling time‑series volatility reveals spikes linked to shocks or seasonal effects, confirming that risk is not constant and requires ongoing monitoring.

**Visual analytics adds measurable value.** Univariate plots highlight skewness and outliers, bivariate analyses reveal correlations between premiums, incurred costs, and claims, while statistical tests validate which categorical differences are meaningful.

**Conclusion:**\
The investigation confirms that profitability and volatility in motor insurance portfolios are primarily shaped by claims frequency and severity, with policy type and bonus score acting as key structural differentiators. Geography and fuel type are less influential, while portfolio risk fluctuates over time. By combining visual analytics with statistical validation, insurers can move beyond descriptive reporting to actionable insights—targeting unstable product types, managing claim drivers, leveraging bonus scores for segmentation, and monitoring volatility trends to anticipate shocks.
:::

```         
::: Speaker Notes

:::
```
