---
title: "Take-home Exercise 1"
author: "Jiang Yuxi"
date: "May 22, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# 1 Overview

## 1.1 Background

Using a real-world open-access motor vehicle insurance portfolio dataset and appropriate visual analytics libraries in R, this applied visual anaytics investigation requires you to demonstrate the potential value of visual analytics techniques for exploring and analysing numerical, categorical, and temporal variables. The objective is to gain deeper insights into data distributions and identifying key factors influencing the profitability and volatility of the insurance portfolio.

## 1.2 The Task

The specific tasks of this study are:

Apply suitable univariate visual analytics techniques to explore the distributions and characteristics of numerical and categorical variables.

Apply suitable bivariate visual analytics techniques to investigate and statistically validate factors influencing the profitability and volatility of the insurance portfolio.

# 2 Getting Started

## 2.1 Loading Packages

```{r}
pacman::p_load(tidyverse, readr, dplyr, patchwork, zoo,
               ggthemes, hrbrthemes, ggridges, plotly,
               forcats, skimr, scales, corrplot, ggiraph) 
```

## 2.2 Data Sources

A real world motor vehicle insurance portfolio data set from a Spanish insurance company specialised in non-life insurance will be used in this study. The data set and its cookbook are available for download at Mendeley Data.

# 3 Data Wrangling

## 3.1 Importing and Inspecting Data

```{r}
portfolio <- read_delim("data/Dataset of motor insurance portfolio.csv", delim = ";")
glimpse(portfolio)
```

### 3.2 Data Type Conversion

Convert categorical variables (policy_type, fuel_type, etc.) into factors.

| Variable | Data Type | Reason |
|---------------------|-------------------|--------------------------------|
| insured_id | integer | Unique identifier, sequential ID. |
| year | integer | Calendar year, discrete not continuous. |
| policy_type | factor | Categorical (COMP_E, COMP_N, TP, etc.). |
| policy_status | factor | Categorical (A = active, C = cancelled). |
| business_type | factor | Categorical (NB = new business, P = portfolio). |
| payment_frequency | factor | Categorical (A, S, Q). |
| bonus_score | factor | Categorical (G, N, B). |
| driver_age | integer | Age in years. |
| vehicle_age | integer | Age in years. |
| age_driving_licence | integer | Year of licence issuance (discrete year). |
| fuel_type | factor | Categorical (D = diesel, G = gasoline). |
| vehicle_value | numeric | Continuous monetary value. |
| seats | integer | Count of seats. |
| power_to_weight_ratio | numeric | Continuous ratio. |
| vehicle_brand | factor | Categorical brand names. |
| municipality_type | factor | Categorical (I, C, IS). |
| circulation_area | factor | Categorical (U, R). |
| \**Premiums (all \_premium)* | numeric | Continuous monetary values. |
| \**Claims (all \_claims)* | integer | Count of claims (discrete). |
| \**Incurred (all \_incurred)* | numeric | Continuous monetary values. |
| \**Exposure (all \_exposure)* | numeric | Continuous, fractional values (0–1 or \>1). |

```{r}
portfolio <- portfolio %>%
  mutate(
    # columns → integer
    insured_id = as.integer(insured_id),
    year = as.integer(year),
    driver_age = as.integer(driver_age),
    vehicle_age = as.integer(vehicle_age),
    age_driving_licence = as.integer(age_driving_licence),
    seats = as.integer(seats),
    total_claims = as.integer(total_claims),
    liability_claims = as.integer(liability_claims),
    liability_property_claims = as.integer(liability_property_claims),
    liability_injury_claims = as.integer(liability_injury_claims),
    property_claims = as.integer(property_claims),
    theft_claims = as.integer(theft_claims),
    fire_claims = as.integer(fire_claims),
    glass_claims = as.integer(glass_claims),
    legal_protection_claims = as.integer(legal_protection_claims),
    occupants_claims = as.integer(occupants_claims),
    
    # columns → factor
    policy_type = factor(policy_type),
    policy_status = factor(policy_status),
    business_type = factor(business_type),
    payment_frequency = factor(payment_frequency),
    bonus_score = factor(bonus_score),
    vehicle_brand = factor(vehicle_brand),
    fuel_type = factor(fuel_type),
    municipality_type = factor(municipality_type),
    circulation_area = factor(circulation_area)
  )
```

### 3.3 Check Consistency

Ensure numeric columns (e.g., vehicle_value, total_premium) are properly parsed as numeric.

```{r}
sapply(portfolio, class)
```

## 3.4 Recoding New Columns

To enhance analytical capability and support profitability and risk analysis, several derived variables were created.

**Profitability Measures** - These are key indicators for evaluating insurance performance.

loss_ratio = total_incurred / total_premium

profit = total_premium - total_incurred

**Risk Metrics** - These measures separate claim occurrence from claim magnitude.

claim_frequency = total_claims / total_exposure

claim_severity = total_incurred / total_claims (for policies with claims)

**Exposure-adjusted Metrics** - Standardizing by exposure ensures fair comparison across policies.

premium_per_exposure

incurred_per_exposure

**Binary Indicators**

has_claim

high_cost_claim

profitable_policy

```{r}
# Precompute threshold for high-cost claims
threshold_incurred <- quantile(portfolio$total_incurred, 0.95, na.rm = TRUE)

portfolio <- portfolio %>%
  mutate(
    # Profitability measures
    loss_ratio = ifelse(total_premium > 0, total_incurred / total_premium, NA_real_),
    profit = total_premium - total_incurred,
    
    # Risk metrics
    claim_frequency = ifelse(total_exposure > 0, total_claims / total_exposure, 0),  # set 0 if exposure=0
    claim_severity = ifelse(total_claims > 0, total_incurred / total_claims, NA_real_),
    
    # Exposure-adjusted metrics
    premium_per_exposure = ifelse(total_exposure > 0, total_premium / total_exposure, NA_real_),
    incurred_per_exposure = ifelse(total_exposure > 0, total_incurred / total_exposure, NA_real_),
    
    # Binary indicators as factors
    has_claim = factor(ifelse(total_claims > 0, 1, 0), levels = c(0,1), labels = c("No Claim","Has Claim")),
    high_cost_claim = factor(ifelse(total_incurred > threshold_incurred, 1, 0), levels = c(0,1), labels = c("Normal","High Cost")),
    profitable_policy = factor(ifelse(profit > 0, 1, 0), levels = c(0,1), labels = c("Not Profitable","Profitable"))
  )
```

### 3.5 Check for Duplicates and Missing values

The below code chunk is used to check whether an entire row (all columns) is repeated.

```{r}
portfolio[duplicated(portfolio),]
```

No rows in this dataset are exact duplicates across all 47 columns.

```{r}
colSums(is.na(portfolio))
```

Vehicle age & Age driving licence → Only 2 missing each → you can safely drop those rows

Fuel type: 1,287 missing (\~0.36% of dataset) → create a new category "Unknown" to preserve those rows.

Vehicle value: 513 missing (\~0.15% of dataset) → use brand-specific median (e.g., median value per vehicle_brand)

Loss ratio → NA when total_premium = 0 (division undefined).

Claim severity → NA when total_claims = 0 (no claims to divide by).

Exposure-adjusted metrics → NA when total_exposure = 0.

```{r}
portfolio <- portfolio %>%
  mutate(
    # Impute vehicle_value with brand-specific median
    vehicle_value = ifelse(is.na(vehicle_value),
                           ave(vehicle_value, vehicle_brand,
                               FUN = function(x) median(x, na.rm = TRUE)),
                           vehicle_value),
    
    # Impute fuel_type with "Unknown"
    fuel_type = fct_explicit_na(factor(fuel_type), na_level = "Unknown")
  )
    
# Drop rows with missing driver/vehicle age
portfolio <- portfolio %>%
  filter(!is.na(vehicle_age), !is.na(age_driving_licence))
```

```{r}
portfolio <- portfolio %>%
  mutate(
    # Flags for structural NA cases
    has_premium = total_premium > 0,
    has_exposure = total_exposure > 0,
    has_claims = total_claims > 0
  )
```

## 3.6 Check for Outliers

```{r}
skim(portfolio)
```

```{r}
# Step 1: Compute IQR thresholds for loss_ratio
Q1_lr <- quantile(portfolio$loss_ratio, 0.25, na.rm = TRUE)
Q3_lr <- quantile(portfolio$loss_ratio, 0.75, na.rm = TRUE)
IQR_lr <- Q3_lr - Q1_lr

# Step 2: Compute IQR thresholds for profit
Q1_pf <- quantile(portfolio$profit, 0.25, na.rm = TRUE)
Q3_pf <- quantile(portfolio$profit, 0.75, na.rm = TRUE)
IQR_pf <- Q3_pf - Q1_pf

# Step 3: Flag outliers for these two columns
portfolio <- portfolio %>%
  mutate(
    is_outlier_loss_ratio = loss_ratio < (Q1_lr - 1.5 * IQR_lr) | loss_ratio > (Q3_lr + 1.5 * IQR_lr),
    is_outlier_profit     = profit < (Q1_pf - 1.5 * IQR_pf) | profit > (Q3_pf + 1.5 * IQR_pf),
    is_outlier            = is_outlier_loss_ratio | is_outlier_profit
  )

# Step 4: Drop rows flagged as outliers
portfolio_drop <- portfolio %>%
  filter(!is_outlier)
```


## 3.7 Recoding Category

```{r}
portfolio <- portfolio %>%
  mutate(
    policy_status = recode(policy_status,
      "A" = "Active",
      "C" = "Cancelled"
    ),

    business_type = recode(business_type,
      "NB" = "New Business",
      "P" = "Portfolio"
    ),

    payment_frequency = recode(payment_frequency,
      "S" = "Semi-annual",
      "Q" = "Quarterly",
      "A" = "Annual"
    ),

    bonus_score = recode(bonus_score,
      "G" = "Good",
      "N" = "Neutral",
      "B" = "Bad"
    ),

    fuel_type = recode(fuel_type,
      "D" = "Diesel",
      "G" = "Gasoline"
    ),

    municipality_type = recode(municipality_type,
      "IS" = "Islands",
      "I" = "Inland",
      "C" = "Coastal"
    ),

    circulation_area = recode(circulation_area,
      "U" = "Urban",
      "R" = "Rural"
    )
  )
```

```{r}
glimpse(portfolio)
```

# 4 Univariate Analysis

Apply suitable univariate visual analytics techniques to explore the distributions and characteristics of numerical and categorical variables.

## 4.1 Cost Distribution 

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
# Reshape incurred cost data into long format
incurred_clean <- portfolio %>%
  select(
    liability_incurred,
    liability_property_incurred,
    liability_injury_incurred,
    property_incurred,
    theft_incurred,
    fire_incurred,
    glass_incurred,
    legal_protection_incurred,
    occupants_incurred
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "claim_type",
    values_to = "cost"
  ) %>%
  filter(is.finite(cost), cost > 0)

# ECDF transformation per claim type
cdf_data <- incurred_clean %>%
  group_by(claim_type) %>%
  arrange(cost) %>%
  mutate(
    cum_pct = row_number() / n() * 100
  ) %>%
  ungroup()

# Plot cumulative distribution
ggplot(cdf_data, aes(x = cum_pct, y = cost, color = claim_type)) +
  geom_line(linewidth = 1, alpha = 0.8) +
  scale_y_log10(
    labels = scales::comma,
    limits = c(1.0, NA) ) +
  labs(
    title = "Cumulative Distribution of Incurred Costs by Claim Type",
    x = "Cumulative Percentage of Claims (%)",
    y = "Incurred Cost (log scale)",
    color = "Claim Type"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold")
  )
```

## The code
::: {style="font-size: 1.4em"}

```{r}
#| eval: false
# Reshape incurred cost data into long format
incurred_clean <- portfolio %>%
  select(
    liability_incurred,
    liability_property_incurred,
    liability_injury_incurred,
    property_incurred,
    theft_incurred,
    fire_incurred,
    glass_incurred,
    legal_protection_incurred,
    occupants_incurred
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "claim_type",
    values_to = "cost"
  ) %>%
  filter(is.finite(cost), cost > 0)

# ECDF transformation per claim type
cdf_data <- incurred_clean %>%
  group_by(claim_type) %>%
  arrange(cost) %>%
  mutate(
    cum_pct = row_number() / n() * 100
  ) %>%
  ungroup()

# Plot cumulative distribution
ggplot(cdf_data, aes(x = cum_pct, y = cost, color = claim_type)) +
  geom_line(linewidth = 1, alpha = 0.8) +
  scale_y_log10(
    labels = scales::comma,
    limits = c(1.0, NA) ) +
  labs(
    title = "Cumulative Distribution of Incurred Costs by Claim Type",
    x = "Cumulative Percentage of Claims (%)",
    y = "Incurred Cost (log scale)",
    color = "Claim Type"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold")
  )
```
:::
::::

## 4.2 Premium Distribution

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
# Reshape incurred cost data into long format
incurred_long <- portfolio %>%
  select(
    liability_premium,
    property_damage_premium,
    theft_premium,
    fire_premium,
    glass_premium,
    legal_protection_premium,
    occupants_premium
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "insurance_type",
    values_to = "premium"
  )

# Stacked density plot
ggplot(incurred_long %>% filter(!is.na(premium), premium > 0),
       aes(x = premium, fill = insurance_type)) +
  geom_density(alpha = 0.6, position = "stack") +
  scale_x_log10(labels = comma) +   # log scale to handle skewness
  theme_minimal() +
  labs(
    title = "Stacked Density Plot of Premium",
    x = "Premuim (log scale)",
    y = "Density"
  ) +
  theme(
    strip.text = element_text(face = "bold"),
    legend.title = element_blank()
  )
```

## The code
::: {style="font-size: 1.4em"}

```{r}
#| eval: false
# Reshape incurred cost data into long format
incurred_long <- portfolio %>%
  select(
    liability_premium,
    property_damage_premium,
    theft_premium,
    fire_premium,
    glass_premium,
    legal_protection_premium,
    occupants_premium
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "insurance_type",
    values_to = "premium"
  )

# Stacked density plot
ggplot(incurred_long %>% filter(!is.na(premium), premium > 0),
       aes(x = premium, fill = insurance_type)) +
  geom_density(alpha = 0.6, position = "stack") +
  scale_x_log10(labels = comma) +   # log scale to handle skewness
  theme_minimal() +
  labs(
    title = "Stacked Density Plot of Premium",
    x = "Premuim (log scale)",
    y = "Density"
  ) +
  theme(
    strip.text = element_text(face = "bold"),
    legend.title = element_blank()
  )
```

:::
::::

## 4.3 Claim Distribution


:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
# Reshape claim count columns into long format
claims_long <- portfolio %>%
  select(
    liability_claims,
    liability_property_claims,
    liability_injury_claims,
    property_claims,
    theft_claims,
    fire_claims,
    glass_claims,
    legal_protection_claims,
    occupants_claims
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "claim_type",
    values_to = "claim_count"
  ) %>%
  filter(is.finite(claim_count))

# Plot boxplots
ggplot(claims_long, aes(x = claim_type, y = claim_count, fill = claim_type)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.4) +
  labs(
    title = "Distribution of Claim Counts by Type",
    x = "Claim Type",
    y = "Number of Claims"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold")
  )
```

## The code
::: {style="font-size: 1.4em"}

```{r}
#| eval: false
# Reshape claim count columns into long format
claims_long <- portfolio %>%
  select(
    liability_claims,
    liability_property_claims,
    liability_injury_claims,
    property_claims,
    theft_claims,
    fire_claims,
    glass_claims,
    legal_protection_claims,
    occupants_claims
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "claim_type",
    values_to = "claim_count"
  ) %>%
  filter(is.finite(claim_count))

# Plot boxplots
ggplot(claims_long, aes(x = claim_type, y = claim_count, fill = claim_type)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.4) +
  labs(
    title = "Distribution of Claim Counts by Type",
    x = "Claim Type",
    y = "Number of Claims"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold")
  )
```
:::
::::

## 4.4 Profitability Metrics

loss_ratio = total_incurred / total_premium  
→ Shows what fraction of premium income is consumed by claims. A ratio > 1 means claims exceed premiums (unprofitable).

profit = total_premium - total_incurred  
→ The net financial result per policy. Positive = profitable, negative = loss.

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
portfolio_bin <- portfolio %>%
  filter(has_premium) %>%
  mutate(
    # Loss ratio bins
    loss_ratio_bin = case_when(
      loss_ratio == 0 ~ "=0",
      loss_ratio > 0 & loss_ratio <= 1 ~ "0-1",
      loss_ratio > 1 ~ ">1",
      TRUE ~ NA_character_
    ),
    # Profit bins
    profit_bin = case_when(
      profit < 0 ~ "Negative",
      profit >= 0 & profit < 250 ~ "0-250",
      profit >= 250 & profit < 500 ~ "250-500",
      profit >= 500 ~ ">500",
      TRUE ~ NA_character_
    )
  )

p1 <- ggplot(portfolio_bin, aes(x = loss_ratio_bin, fill = loss_ratio_bin)) +
  geom_bar(alpha = 0.7) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  labs(title = "Loss Ratio Bins", x = "Loss Ratio Category", y = "Count") +
  scale_x_discrete(drop = TRUE) +
  theme_minimal() +
  theme(legend.position = "none")

p2 <- ggplot(portfolio_bin, aes(x = profit_bin, fill = profit_bin)) +
  geom_bar(alpha = 0.7) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  labs(title = "Profit Bins", x = "Profit Category", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")

p1 + p2
```

## The code
::: {style="font-size: 1.4em"}

```{r}
#| eval: false
portfolio_bin <- portfolio %>%
  filter(has_premium) %>%
  mutate(
    # Loss ratio bins
    loss_ratio_bin = case_when(
      loss_ratio == 0 ~ "=0",
      loss_ratio > 0 & loss_ratio <= 1 ~ "0-1",
      loss_ratio > 1 ~ ">1",
      TRUE ~ NA_character_
    ),
    # Profit bins
    profit_bin = case_when(
      profit < 0 ~ "Negative",
      profit >= 0 & profit < 250 ~ "0-250",
      profit >= 250 & profit < 500 ~ "250-500",
      profit >= 500 ~ ">500",
      TRUE ~ NA_character_
    )
  )

p1 <- ggplot(portfolio_bin, aes(x = loss_ratio_bin, fill = loss_ratio_bin)) +
  geom_bar(alpha = 0.7) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  labs(title = "Loss Ratio Bins", x = "Loss Ratio Category", y = "Count") +
  scale_x_discrete(drop = TRUE) +
  theme_minimal() +
  theme(legend.position = "none")

p2 <- ggplot(portfolio_bin, aes(x = profit_bin, fill = profit_bin)) +
  geom_bar(alpha = 0.7) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  labs(title = "Profit Bins", x = "Profit Category", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")

p1 + p2
```

:::
::::

## 4.5 Categorical variables

Bar charts → for policy_type, fuel_type, municipality_type, circulation_area, bonus_score.

Pie charts (optional) → for simple categorical distributions like policy_status.

Frequency tables → to confirm counts and proportions.

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
# Categorical variables: bar charts
# Select categorical variables
cat_long <- portfolio %>%
  select(
    policy_type,
    policy_status,
    business_type,
    payment_frequency,
    bonus_score,
    fuel_type,
    municipality_type,
    circulation_area
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  )

# Create percentage table within each variable
plot_data <- cat_long %>%
  group_by(variable, value) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(variable) %>%
  mutate(
    percent = count / sum(count),
    value = fct_reorder(value, percent)
  )

# Plot
ggplot(plot_data,
       aes(x = percent,
           y = value)) +
  geom_col(fill = "steelblue") +
  facet_wrap(~variable,
             scales = "free_y") +
  scale_x_continuous(labels = percent_format()) +
  labs(
    title = "Distribution of Categorical Variables",
    x = "Percentage",
    y = NULL
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold"),
    axis.text.y = element_text(size = 9)
  )
```

## The code
::: {style="font-size: 1.4em"}

```{r}
#| eval: false
# Categorical variables: bar charts
# Select categorical variables
cat_long <- portfolio %>%
  select(
    policy_type,
    policy_status,
    business_type,
    payment_frequency,
    bonus_score,
    fuel_type,
    municipality_type,
    circulation_area
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  )

# Create percentage table within each variable
plot_data <- cat_long %>%
  group_by(variable, value) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(variable) %>%
  mutate(
    percent = count / sum(count),
    value = fct_reorder(value, percent)
  )

# Plot
ggplot(plot_data,
       aes(x = percent,
           y = value)) +
  geom_col(fill = "steelblue") +
  facet_wrap(~variable,
             scales = "free_y") +
  scale_x_continuous(labels = percent_format()) +
  labs(
    title = "Distribution of Categorical Variables",
    x = "Percentage",
    y = NULL
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold"),
    axis.text.y = element_text(size = 9)
  )
```

:::
::::

# 5 Exploratory Data Analysis

Apply suitable bivariate visual analytics techniques to investigate and statistically validate factors influencing the profitability and volatility of the insurance portfolio.

### 5.1 Correlation Analysis

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
# Select numeric variables
num_vars <- portfolio %>%
  select(total_premium, total_incurred, profit, loss_ratio,
         claim_frequency, claim_severity, total_claims,
         premium_per_exposure, incurred_per_exposure,
         driver_age, vehicle_age, power_to_weight_ratio, vehicle_value)

# Correlation matrix
cor_matrix <- cor(num_vars, use = "complete.obs", method = "spearman")

# Visualize
corrplot.mixed(cor_matrix,
               lower = "ellipse",
               upper = "number",
               number.cex = 0.5,
               tl.cex = 0.8,
               order = "hclust",
               tl.pos = "lt",
               diag = "l",
               tl.col = "black",
               na.label = " ")   

# Example correlation test
cor.test(portfolio$claim_frequency, portfolio$loss_ratio, method = "spearman")
```

## The code
::: {style="font-size: 1.4em"}

```{r}
#| eval: false
# Select numeric variables
num_vars <- portfolio %>%
  select(total_premium, total_incurred, profit, loss_ratio,
         claim_frequency, claim_severity, total_claims,
         premium_per_exposure, incurred_per_exposure,
         driver_age, vehicle_age, power_to_weight_ratio, vehicle_value)

# Correlation matrix
cor_matrix <- cor(num_vars, use = "complete.obs", method = "spearman")

# Visualize
corrplot.mixed(cor_matrix,
               lower = "ellipse",
               upper = "number",
               number.cex = 0.5,
               tl.cex = 0.8,
               order = "hclust",
               tl.pos = "lt",
               diag = "l",
               tl.col = "black",
               na.label = " ")   

# Example correlation test
cor.test(portfolio$claim_frequency, portfolio$loss_ratio, method = "spearman")
```
:::
::::

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
cor_df <- cor_matrix %>%
  as.data.frame() %>%
  rownames_to_column("var1") %>%
  pivot_longer(-var1, names_to = "var2", values_to = "correlation") %>%
  filter(var1 < var2, abs(correlation) >= 0.5) %>%
  arrange(desc(abs(correlation)))

print(cor_df)
```

## The code
::: {style="font-size: 1.4em"}

```{r}
#| eval: false
cor_df <- cor_matrix %>%
  as.data.frame() %>%
  rownames_to_column("var1") %>%
  pivot_longer(-var1, names_to = "var2", values_to = "correlation") %>%
  filter(var1 < var2, abs(correlation) >= 0.5) %>%
  arrange(desc(abs(correlation)))

print(cor_df)
```
:::
::::

::: {.callout-note}
## Insight:

Profitability is dominated by loss ratio → controlling incurred relative to premium is key.

Volatility is driven more by claim severity than frequency → catastrophic claims destabilize the portfolio.

Exposure-adjusted metrics (premium/incurred per exposure) are excellent volatility indicators.

Vehicle and driver demographics show moderate effects, but not as strong as financial ratios.
:::

## 5.2 Profitability vs Risk Metrics

### 5.2.1 Premium vs Incurred

Scatter plot with smoothing line to see how incurred costs relate to premiums, highlighting profitability patterns.

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
# Premium vs Incurred (profitability)

portfolio_clean <- portfolio %>%
  filter(total_incurred > 0, total_premium >= 0)

ggplot(portfolio, aes(x = total_premium, y = total_incurred)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  scale_y_log10() +
  coord_cartesian(xlim = c(0, 3000), ylim = c(1, 500000)) +  
  geom_smooth(method = "lm", se = TRUE, color = "darkred") +
  labs(
    title = "Premium vs Incurred Costs",
    x = "Total Premium",
    y = "Total Incurred"
  ) +
  theme_minimal(base_size = 14)
```

## The code
::: {style="font-size: 1.4em"}

```{r}
#| eval: false
portfolio_clean <- portfolio %>%
  filter(total_incurred > 0, total_premium >= 0)

ggplot(portfolio, aes(x = total_premium, y = total_incurred)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  scale_y_log10() +
  coord_cartesian(xlim = c(0, 3000), ylim = c(1, 500000)) +  
  geom_smooth(method = "lm", se = TRUE, color = "darkred") +
  labs(
    title = "Premium vs Incurred Costs",
    x = "Total Premium",
    y = "Total Incurred"
  ) +
  theme_minimal(base_size = 14)
```

:::
::::


### 5.2.2 Loss Ratio vs Claim frequency

Scatter plot with smoothing line to see if higher claim frequency drives higher loss ratios. The regression line gives a quick statistical validation.

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
ggplot(portfolio %>% filter(has_premium), aes(x = claim_frequency, y = loss_ratio)) +
  geom_point(alpha = 0.4, color = "steelblue") +   # scatter points
  coord_cartesian(xlim = c(0, 300), ylim = c(0, 3000)) +
  geom_smooth(method = "lm", se = TRUE, color = "darkred") +  # regression line
  labs(
    title = "Loss Ratio vs Claim Frequency",
    x = "Claim Frequency (claims per exposure)",
    y = "Loss Ratio"
  ) +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"))
```

## The code
::: {style="font-size: 1.4em"}

```{r}
#| eval: false
ggplot(portfolio  %>% filter(has_premium), aes(x = claim_frequency, y = loss_ratio)) +
  geom_point(alpha = 0.4, color = "steelblue") +   # scatter points
  coord_cartesian(xlim = c(0, 300), ylim = c(0, 3000)) +
  geom_smooth(method = "lm", se = TRUE, color = "darkred") +  # regression line
  labs(
    title = "Loss Ratio vs Claim Frequency",
    x = "Claim Frequency (claims per exposure)",
    y = "Loss Ratio"
  ) +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"))
```

:::
::::

### 5.2.3 Profit vs Claim Severity

Boxplots grouped by severity bins to check if extreme claims erode profitability.

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
# Step 1: Create severity bins
portfolio_sev <- portfolio %>%
  mutate(
    severity_bin = case_when(
      claim_severity < 1000 ~ "<1k",
      claim_severity >= 1000 & claim_severity < 5000 ~ "1k-5k",
      claim_severity >= 5000 & claim_severity < 20000 ~ "5k-20k",
      claim_severity >= 20000 ~ ">20k",
      TRUE ~ NA_character_
    )
  )

# Step 2: Compute median profit per severity bin
summary_stats <- portfolio_sev %>%
  filter(!is.na(severity_bin)) %>%
  group_by(severity_bin) %>%
  summarise(median_profit = median(profit, na.rm = TRUE))

# Step 3: Boxplot with summary labels
ggplot(portfolio_sev %>% filter(!is.na(severity_bin)),
       aes(x = severity_bin, y = profit, fill = severity_bin)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  coord_cartesian(ylim = c(-400000 , 200000 )) +
  geom_text(data = summary_stats,
            aes(x = severity_bin, y = median_profit,
                label = scales::comma(round(median_profit, 0))),
            vjust = -0.5, fontface = "bold", color = "darkred") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Profit vs Claim Severity",
    x = "Claim Severity Bin",
    y = "Profit"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold")
  )
```

## The code
::: {style="font-size: 1.4em"}

```{r}
#| eval: false
# Step 1: Create severity bins
portfolio_sev <- portfolio %>%
  mutate(
    severity_bin = case_when(
      claim_severity < 1000 ~ "<1k",
      claim_severity >= 1000 & claim_severity < 5000 ~ "1k-5k",
      claim_severity >= 5000 & claim_severity < 20000 ~ "5k-20k",
      claim_severity >= 20000 ~ ">20k",
      TRUE ~ NA_character_
    )
  )

# Step 2: Compute median profit per severity bin
summary_stats <- portfolio_sev %>%
  filter(!is.na(severity_bin)) %>%
  group_by(severity_bin) %>%
  summarise(median_profit = median(profit, na.rm = TRUE))

# Step 3: Boxplot with summary labels
ggplot(portfolio_sev %>% filter(!is.na(severity_bin)),
       aes(x = severity_bin, y = profit, fill = severity_bin)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  coord_cartesian(ylim = c(-400000 , 200000 )) +
  geom_text(data = summary_stats,
            aes(x = severity_bin, y = median_profit,
                label = scales::comma(round(median_profit, 0))),
            vjust = -0.5, fontface = "bold", color = "darkred") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Profit vs Claim Severity",
    x = "Claim Severity Bin",
    y = "Profit"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold")
  )
```
:::
::::


## 5.3 Policy & Vehicle Characteristics

### 5.3.1 Policy & Vehicle Characteristics

Bar chart showing mean profit per policy type, with confidence intervals. This shows average profitability across policy types (e.g., COMP_E, COMP_N, TP), with bootstrapped confidence intervals.

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
# Mean profit per policy type with confidence intervals
ggplot(portfolio, aes(x = policy_type, y = profit, fill = policy_type)) +
  stat_summary(fun = mean, geom = "bar", alpha = 0.7) +
  stat_summary(fun.data = mean_cl_boot, geom = "errorbar", width = 0.2) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Mean Profit by Policy Type",
    x = "Policy Type",
    y = "Mean Profit"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold")
  )
```

## The code
::: {style="font-size: 1.4em"}

```{r}
#| eval: false
ggplot(portfolio, aes(x = policy_type, y = profit, fill = policy_type)) +
  stat_summary(fun = mean, geom = "bar", alpha = 0.7) +
  stat_summary(fun.data = mean_cl_boot, geom = "errorbar", width = 0.2) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Mean Profit by Policy Type",
    x = "Policy Type",
    y = "Mean Profit"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold")
  )
```

:::
::::

### 5.3.2 Fuel type vs loss ratio

Violin plots to compare distributions across Diesel, Gasoline, Unknown. Violin plots show the full distribution shape, while the embedded boxplot highlights median and quartiles.

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
# Violin plot: Loss ratio by fuel type
ggplot(portfolio %>% filter(has_premium), aes(x = fuel_type, y = loss_ratio, fill = fuel_type)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  coord_cartesian(ylim = c(0 , 5000)) +
  geom_boxplot(width = 0.1, fill = "white", outlier.alpha = 0.3) +
  labs(
    title = "Loss Ratio Distribution by Fuel Type",
    x = "Fuel Type",
    y = "Loss Ratio"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold")
  )
```

## The code
::: {style="font-size: 1.4em"}

```{r}
#| eval: false
ggplot(portfolio %>% filter(has_premium), aes(x = fuel_type, y = loss_ratio, fill = fuel_type)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  coord_cartesian(ylim = c(0 , 5000)) +
  geom_boxplot(width = 0.1, fill = "white", outlier.alpha = 0.3) +
  labs(
    title = "Loss Ratio Distribution by Fuel Type",
    x = "Fuel Type",
    y = "Loss Ratio"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold")
  )
```
:::
::::

### 5.3.3 Vehicle Age vs Claim Frequency

Scatter plot with regression line to test if older vehicles have higher claim frequency. This reveals whether older vehicles tend to have higher claim frequency. The regression line provides a statistical trend.

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
# Scatter plot: Vehicle age vs Claim frequency
ggplot(portfolio, aes(x = vehicle_age, y = claim_frequency)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  labs(
    title = "Vehicle Age vs Claim Frequency",
    x = "Vehicle Age (years)",
    y = "Claim Frequency (claims per exposure)"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))
```

## The code
::: {style="font-size: 1.4em"}

```{r}
#| eval: false
ggplot(portfolio, aes(x = vehicle_age, y = claim_frequency)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  labs(
    title = "Vehicle Age vs Claim Frequency",
    x = "Vehicle Age (years)",
    y = "Claim Frequency (claims per exposure)"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))
```
:::
::::

## 5.4 Driver Age vs Loss Ratio

This compares the distribution of loss ratios across age bands. This highlights which age groups tend to have higher or lower loss ratios.

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
# Create age bands
portfolio_age <- portfolio %>% 
  filter(has_premium) %>%
  mutate(
    age_band = cut(driver_age,
                   breaks = seq(18, 80, by = 10),
                   include.lowest = TRUE)
  ) %>%
  group_by(age_band) %>%
  mutate(
    q1 = quantile(loss_ratio, 0.25, na.rm = TRUE),
    q3 = quantile(loss_ratio, 0.75, na.rm = TRUE),
    iqr = q3 - q1,
    lower = q1 - 1.5 * iqr,
    upper = q3 + 1.5 * iqr
  ) %>%
  ungroup() %>%
  filter(loss_ratio >= lower, loss_ratio <= upper)

# Ridge plot of loss ratio by age band
ggplot(portfolio_age, aes(x = loss_ratio, y = age_band, fill = age_band)) +
  ggridges::geom_density_ridges(alpha = 0.6) +
  labs(
    title = "Loss Ratio Distribution by Driver Age Band",
    x = "Loss Ratio",
    y = "Driver Age Band"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## The code
::: {style="font-size: 1.4em"}

```{r}
#| eval: false
# Create age bands
portfolio_age <- portfolio %>% 
  filter(has_premium) %>%
  mutate(
    age_band = cut(driver_age,
                   breaks = seq(18, 80, by = 10),
                   include.lowest = TRUE)
  ) %>%
  group_by(age_band) %>%
  mutate(
    q1 = quantile(loss_ratio, 0.25, na.rm = TRUE),
    q3 = quantile(loss_ratio, 0.75, na.rm = TRUE),
    iqr = q3 - q1,
    lower = q1 - 1.5 * iqr,
    upper = q3 + 1.5 * iqr
  ) %>%
  ungroup() %>%
  filter(loss_ratio >= lower, loss_ratio <= upper)

# Ridge plot of loss ratio by age band
ggplot(portfolio_age, aes(x = loss_ratio, y = age_band, fill = age_band)) +
  ggridges::geom_density_ridges(alpha = 0.6) +
  labs(
    title = "Loss Ratio Distribution by Driver Age Band",
    x = "Loss Ratio",
    y = "Driver Age Band"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
:::
::::

## 5.5 Volatility 

Claims frequency analysis → average claims per policy type, per fuel type, per circulation area.

Severity analysis → mean incurred per claim across categories.

Exposure vs claims → scatter plot or regression to validate exposure as a driver of volatility.

### 5.5.1 Loss ratio variability  

Measures how much the ratio of incurred to premium fluctuates across policies. High variance = unstable underwriting results.

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
# Variance of loss ratio by policy type
portfolio %>%
  group_by(policy_type) %>%
  summarise(
    mean_lr = mean(loss_ratio, na.rm = TRUE),
    sd_lr   = sd(loss_ratio, na.rm = TRUE),
    var_lr  = var(loss_ratio, na.rm = TRUE)
  )
```

## The code
::: {style="font-size: 1.4em"}

```{r}
#| eval: false
portfolio %>%
  group_by(policy_type) %>%
  summarise(
    mean_lr = mean(loss_ratio, na.rm = TRUE),
    sd_lr   = sd(loss_ratio, na.rm = TRUE),
    var_lr  = var(loss_ratio, na.rm = TRUE)
  )
```
:::
::::

### 5.5.2 Coefficient of Variation 

CV = sd/mean, useful for claim frequency and severity because it normalizes volatility relative to the mean.

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
# CV for claim frequency and severity
portfolio %>%
  summarise(
    mean_freq = mean(claim_frequency, na.rm = TRUE),
    sd_freq   = sd(claim_frequency, na.rm = TRUE),
    cv_freq   = sd_freq / mean_freq,
    
    mean_sev = mean(claim_severity, na.rm = TRUE),
    sd_sev   = sd(claim_severity, na.rm = TRUE),
    cv_sev   = sd_sev / mean_sev
  )
```

## The code
::: {style="font-size: 1.4em"}

```{r}
#| eval: false
portfolio %>%
  summarise(
    mean_freq = mean(claim_frequency, na.rm = TRUE),
    sd_freq   = sd(claim_frequency, na.rm = TRUE),
    cv_freq   = sd_freq / mean_freq,
    
    mean_sev = mean(claim_severity, na.rm = TRUE),
    sd_sev   = sd(claim_severity, na.rm = TRUE),
    cv_sev   = sd_sev / mean_sev
  )
```
:::
::::

### 5.5.3 Levene's Test

Statistically validate whether variance differs across groups (e.g., urban vs rural, diesel vs gasoline).

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
library(car)

# Levene’s test for loss ratio variance across circulation areas
leveneTest(loss_ratio ~ circulation_area, data = portfolio)

# Levene’s test for claim severity variance across fuel types
leveneTest(claim_severity ~ fuel_type, data = portfolio)
```

## The code
::: {style="font-size: 1.4em"}

```{r}
#| eval: false
# Levene’s test for loss ratio variance across circulation areas
leveneTest(loss_ratio ~ circulation_area, data = portfolio)

# Levene’s test for claim severity variance across fuel types
leveneTest(claim_severity ~ fuel_type, data = portfolio)
```
:::
::::

### 5.5.4 Time-series Volatility

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
library(zoo)
portfolio_ts <- portfolio %>%
  group_by(year) %>%
  summarise(ratio = mean(total_incurred / total_premium, na.rm = TRUE))

portfolio_ts$rolling_sd <- rollapply(portfolio_ts$ratio, width = 3, FUN = sd, fill = NA, align = "right")
```

## The code
::: {style="font-size: 1.4em"}

```{r}
#| eval: false
portfolio_ts <- portfolio %>%
  group_by(year) %>%
  summarise(ratio = mean(total_incurred / total_premium, na.rm = TRUE))

portfolio_ts$rolling_sd <- rollapply(portfolio_ts$ratio, width = 3, FUN = sd, fill = NA, align = "right")
```
:::
::::

## 5.6 Hypothesis Testing


### 5.6.1 Chi-square tests → categorical vs categorical (e.g., policy_type vs policy_status).

Use for categorical vs categorical relationships:

Policy type vs profitability → Is profitability distribution independent of policy type?

Bonus score vs profitability → Are Good/Neutral/Bad scores associated with different profitability outcomes?

Circulation area vs has_claim → Do urban vs rural areas differ in claim occurrence?

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
chisq.test(table(portfolio$policy_type, portfolio$profitable_policy))
chisq.test(table(portfolio$bonus_score, portfolio$profitable_policy))
chisq.test(table(portfolio$circulation_area, portfolio$has_claim))
```

## The code
::: {style="font-size: 1.4em"}

```{r}
#| eval: false
chisq.test(table(portfolio$policy_type, portfolio$profitable_policy))
chisq.test(table(portfolio$bonus_score, portfolio$profitable_policy))
chisq.test(table(portfolio$circulation_area, portfolio$has_claim))
```
:::
::::

### 5.6.2 t-tests / ANOVA → numeric vs categorical (e.g., total_incurred across fuel_type).

Use for continuous vs categorical comparisons:

Loss ratio vs fuel type → Do Diesel, Gasoline, Unknown differ in mean loss ratio?

Profit vs business type → Compare mean profit between New Business vs Portfolio.

Claim severity vs municipality type → Inland vs Coastal vs Islands.

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
anova_result <- aov(loss_ratio ~ fuel_type, data = portfolio)
summary(anova_result)
TukeyHSD(anova_result)
```

## The code
::: {style="font-size: 1.4em"}

```{r}
#| eval: false
anova_result <- aov(loss_ratio ~ fuel_type, data = portfolio)
summary(anova_result)
TukeyHSD(anova_result)
```
:::
::::

# 6 Summary

# 7 References and Inspiration
