---
title: "Take-home Exercise 1"
author: "Jiang Yuxi"
date: "May 22, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Exploring Profitability and Volatility in Motor Insurance

# 1 Overview

## 1.1 Background

Using a real-world open-access motor vehicle insurance portfolio dataset and appropriate visual analytics libraries in R, this applied visual anaytics investigation requires you to demonstrate the potential value of visual analytics techniques for exploring and analysing numerical, categorical, and temporal variables. The objective is to gain deeper insights into data distributions and identifying key factors influencing the profitability and volatility of the insurance portfolio.

## 1.2 The Task

The specific tasks of this study are:

Apply suitable univariate visual analytics techniques to explore the distributions and characteristics of numerical and categorical variables.

Apply suitable bivariate visual analytics techniques to investigate and statistically validate factors influencing the profitability and volatility of the insurance portfolio.

# 2 Getting Started

## 2.1 Loading Packages

| **Package** | **Category** | **Purpose / Usage** |
|--------------------------|-------------------------------------|----------|
| **tidyverse** | Data wrangling + visualization | Core suite (ggplot2 for plots, dplyr for mutate, summarise, group_by). |
| **readr** | Data import | Reading CSV/flat files. |
| **dplyr** | Data manipulation | Filtering, mutating, summarising, grouping. |
| **patchwork** | Visualization | Combining multiple ggplot2 plots into composite layouts. |
| **zoo** | Time series | `rollapply()` for rolling volatility in time‑series analysis. |
| **ggridges** | Visualization | Ridge plots for driver age band distributions. |
| **plotly** | Interactive visualization | Interactive plots (hover, zoom, tooltips). |
| **forcats** | Factor handling | Reordering and recoding categorical variables. |
| **skimr** | Data diagnostics | Quick dataset summaries (missing values, distributions). |
| **scales** | Visualization | Formatting axis labels (comma, percent, currency). |
| **corrplot** | Visualization | Correlation matrix visualization. |
| **ggiraph** | Interactive visualization | Interactive ggplot2 graphics (clickable, hoverable elements). |
| **car** | Statistical testing | `leveneTest()` for variance homogeneity tests. |

pacman::p_load(...) is used to load all the required R libraries for data wrangling, visualization, and statistical analysis in one go.

```{r}
pacman::p_load(tidyverse, readr, dplyr, patchwork, zoo, ggridges, plotly,
               forcats, skimr, scales, corrplot, ggiraph) 
```

## 2.2 Data Sources

Motor vehicle insurance portfolio data set from a Spanish insurance company specialised in non-life insurance will be used in this study. The data set and its cookbook are available for download at [Mendeley Data](https://data.mendeley.com/datasets/sw4jmdb2sm/1).

# 3 Data Wrangling

## 3.1 Importing and Inspecting Data

read_delim(...) together with glimpse(portfolio) is used to read the motor insurance dataset from CSV and display its structure for a quick overview.

```{r}
portfolio <- read_delim("data/Dataset of motor insurance portfolio.csv", delim = ";")
glimpse(portfolio)
```

### 3.2 Data Type Conversion

Convert categorical variables (policy_type, fuel_type, etc.) into factors.

| Variable | Data Type | Reason |
|---------------------|-------------------|--------------------------------|
| insured_id | integer | Unique identifier, sequential ID. |
| year | integer | Calendar year, discrete not continuous. |
| policy_type | factor | Categorical (COMP_E, COMP_N, TP, etc.). |
| policy_status | factor | Categorical (A = active, C = cancelled). |
| business_type | factor | Categorical (NB = new business, P = portfolio). |
| payment_frequency | factor | Categorical (A, S, Q). |
| bonus_score | factor | Categorical (G, N, B). |
| driver_age | integer | Age in years. |
| vehicle_age | integer | Age in years. |
| age_driving_licence | integer | Year of licence issuance (discrete year). |
| fuel_type | factor | Categorical (D = diesel, G = gasoline). |
| vehicle_value | numeric | Continuous monetary value. |
| seats | integer | Count of seats. |
| power_to_weight_ratio | numeric | Continuous ratio. |
| vehicle_brand | factor | Categorical brand names. |
| municipality_type | factor | Categorical (I, C, IS). |
| circulation_area | factor | Categorical (U, R). |
| \**Premiums (all \_premium)* | numeric | Continuous monetary values. |
| \**Claims (all \_claims)* | integer | Count of claims (discrete). |
| \**Incurred (all \_incurred)* | numeric | Continuous monetary values. |
| \**Exposure (all \_exposure)* | numeric | Continuous, fractional values (0–1 or \>1). |

mutate(... as.integer / factor) is used to convert variables into the correct formats, such as integers for IDs and factors for categorical fields.

```{r}
portfolio <- portfolio %>%
  mutate(
    # columns → integer
    insured_id = as.integer(insured_id),
    year = as.integer(year),
    driver_age = as.integer(driver_age),
    vehicle_age = as.integer(vehicle_age),
    age_driving_licence = as.integer(age_driving_licence),
    seats = as.integer(seats),
    total_claims = as.integer(total_claims),
    liability_claims = as.integer(liability_claims),
    liability_property_claims = as.integer(liability_property_claims),
    liability_injury_claims = as.integer(liability_injury_claims),
    property_claims = as.integer(property_claims),
    theft_claims = as.integer(theft_claims),
    fire_claims = as.integer(fire_claims),
    glass_claims = as.integer(glass_claims),
    legal_protection_claims = as.integer(legal_protection_claims),
    occupants_claims = as.integer(occupants_claims),
    
    # columns → factor
    policy_type = factor(policy_type),
    policy_status = factor(policy_status),
    business_type = factor(business_type),
    payment_frequency = factor(payment_frequency),
    bonus_score = factor(bonus_score),
    vehicle_brand = factor(vehicle_brand),
    fuel_type = factor(fuel_type),
    municipality_type = factor(municipality_type),
    circulation_area = factor(circulation_area)
  )
```

### 3.3 Check Consistency

sapply(portfolio, class) is used to verify that numeric and categorical columns are correctly parsed.

```{r}
sapply(portfolio, class)
```

## 3.4 Recoding New Columns

To enhance analytical capability and support profitability and risk analysis, several derived variables were created.

**Profitability Measures** - These are key indicators for evaluating insurance performance.

-   loss_ratio = total_incurred / total_premium

-   profit = total_premium - total_incurred

**Risk Metrics** - These measures separate claim occurrence from claim magnitude.

-   claim_frequency = total_claims / total_exposure

-   claim_severity = total_incurred / total_claims (for policies with claims)

**Exposure-adjusted Metrics** - Standardizing by exposure ensures fair comparison across policies.

-   premium_per_exposure = total_premium / total_exposure

-   incurred_per_exposure = total_incurred / total_exposure

**Binary Indicators**

-   has_claim: total_claims \> 0, 1, 0

-   high_cost_claim: total_incurred \> threshold_incurred, 1, 0

-   profitable_policy: profit \> 0, 1, 0

mutate(loss_ratio, profit, claim_frequency, ...) is used to create new profitability and risk metrics such as loss ratio, profit, claim frequency, claim severity, and exposure‑adjusted measures.

```{r}
# Precompute threshold for high-cost claims
threshold_incurred <- quantile(portfolio$total_incurred, 0.95, na.rm = TRUE)

portfolio <- portfolio %>%
  mutate(
    # Profitability measures
    loss_ratio = ifelse(total_premium > 0, total_incurred / total_premium, NA_real_),
    profit = total_premium - total_incurred,
    
    # Risk metrics
    claim_frequency = ifelse(total_exposure > 0, total_claims / total_exposure, 0),  # set 0 if exposure=0
    claim_severity = ifelse(total_claims > 0, total_incurred / total_claims, NA_real_),
    
    # Exposure-adjusted metrics
    premium_per_exposure = ifelse(total_exposure > 0, total_premium / total_exposure, NA_real_),
    incurred_per_exposure = ifelse(total_exposure > 0, total_incurred / total_exposure, NA_real_),
    
    # Binary indicators as factors
    has_claim = factor(ifelse(total_claims > 0, 1, 0), levels = c(0,1), labels = c("No Claim","Has Claim")),
    high_cost_claim = factor(ifelse(total_incurred > threshold_incurred, 1, 0), levels = c(0,1), labels = c("Normal","High Cost")),
    profitable_policy = factor(ifelse(profit > 0, 1, 0), levels = c(0,1), labels = c("Not Profitable","Profitable"))
  )
```

### 3.5 Check for Duplicates and Missing values

duplicated() is used to check whether an entire row (all columns) is repeated.

```{r}
portfolio[duplicated(portfolio),]
```

No rows in this dataset are exact duplicates across all 56 columns.

```{r}
colSums(is.na(portfolio))
```

Vehicle age & Age driving licence → Only 2 missing each → you can safely drop those rows

Fuel type: 1,287 missing (\~0.36% of dataset) → create a new category "Unknown" to preserve those rows.

Vehicle value: 513 missing (\~0.15% of dataset) → use brand-specific median (e.g., median value per vehicle_brand)

Loss ratio → NA when total_premium = 0 (division undefined).

Claim severity → NA when total_claims = 0 (no claims to divide by).

Exposure-adjusted metrics → NA when total_exposure = 0.

colSums(is.na(...)) together with imputation code is used to identify missing data, fill vehicle values using brand‑specific medians, assign “Unknown” to missing fuel types, and drop rows missing driver or vehicle age.

```{r}
portfolio <- portfolio %>%
  mutate(
    # Impute vehicle_value with brand-specific median
    vehicle_value = ifelse(is.na(vehicle_value),
                           ave(vehicle_value, vehicle_brand,
                               FUN = function(x) median(x, na.rm = TRUE)),
                           vehicle_value),
    
    # Impute fuel_type with "Unknown"
    fuel_type = fct_explicit_na(factor(fuel_type), na_level = "Unknown")
  )
    
# Drop rows with missing driver/vehicle age
portfolio <- portfolio %>%
  filter(!is.na(vehicle_age), !is.na(age_driving_licence))
```

```{r}
portfolio <- portfolio %>%
  mutate(
    # Flags for structural NA cases
    has_premium = total_premium > 0,
    has_exposure = total_exposure > 0,
    has_claims = total_claims > 0
  )
```

## 3.6 Check for Outliers

```{r}
skim(portfolio)
```

IQR calculations with mutate(is_outlier...) are used to flag extreme values in loss ratio and profit, and then remove those outlier rows from the dataset.

```{r}
# Step 1: Compute IQR thresholds for loss_ratio
Q1_lr <- quantile(portfolio$loss_ratio, 0.25, na.rm = TRUE)
Q3_lr <- quantile(portfolio$loss_ratio, 0.75, na.rm = TRUE)
IQR_lr <- Q3_lr - Q1_lr

# Step 2: Compute IQR thresholds for profit
Q1_pf <- quantile(portfolio$profit, 0.25, na.rm = TRUE)
Q3_pf <- quantile(portfolio$profit, 0.75, na.rm = TRUE)
IQR_pf <- Q3_pf - Q1_pf

# Step 3: Flag outliers for these two columns
portfolio <- portfolio %>%
  mutate(
    is_outlier_loss_ratio = loss_ratio < (Q1_lr - 1.5 * IQR_lr) | loss_ratio > (Q3_lr + 1.5 * IQR_lr),
    is_outlier_profit     = profit < (Q1_pf - 1.5 * IQR_pf) | profit > (Q3_pf + 1.5 * IQR_pf),
    is_outlier            = is_outlier_loss_ratio | is_outlier_profit
  )

# Step 4: Drop rows flagged as outliers
portfolio_drop <- portfolio %>%
  filter(!is_outlier)
```

## 3.7 Recoding Category

recode(...) is used to replace coded categorical values with descriptive labels, such as “A” becoming “Active” and “D” becoming “Diesel.”

```{r}
portfolio <- portfolio %>%
  mutate(
    policy_status = recode(policy_status,
      "A" = "Active",
      "C" = "Cancelled"
    ),

    business_type = recode(business_type,
      "NB" = "New Business",
      "P" = "Portfolio"
    ),

    payment_frequency = recode(payment_frequency,
      "S" = "Semi-annual",
      "Q" = "Quarterly",
      "A" = "Annual"
    ),

    bonus_score = recode(bonus_score,
      "G" = "Good",
      "N" = "Neutral",
      "B" = "Bad"
    ),

    fuel_type = recode(fuel_type,
      "D" = "Diesel",
      "G" = "Gasoline"
    ),

    municipality_type = recode(municipality_type,
      "IS" = "Islands",
      "I" = "Inland",
      "C" = "Coastal"
    ),

    circulation_area = recode(circulation_area,
      "U" = "Urban",
      "R" = "Rural"
    )
  )
```

```{r}
glimpse(portfolio)
```

# 4 Univariate Analysis

Apply suitable univariate visual analytics techniques to explore the distributions and characteristics of numerical and categorical variables.

## 4.1 Cost Distribution

This code reshapes the incurred cost data into a long format, computes the empirical cumulative distribution (ECDF) for each claim type, and plots cumulative distribution curves. The purpose is to show how incurred costs accumulate across different claim types, highlighting the spread and scale of claim costs in the portfolio.

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
# Reshape incurred cost data into long format
incurred_clean <- portfolio %>%
  select(
    liability_incurred,
    liability_property_incurred,
    liability_injury_incurred,
    property_incurred,
    theft_incurred,
    fire_incurred,
    glass_incurred,
    legal_protection_incurred,
    occupants_incurred
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "claim_type",
    values_to = "cost"
  ) %>%
  filter(is.finite(cost), cost > 0)

# ECDF transformation per claim type
cdf_data <- incurred_clean %>%
  group_by(claim_type) %>%
  arrange(cost) %>%
  mutate(
    cum_pct = row_number() / n() * 100
  ) %>%
  ungroup()

# Plot cumulative distribution
ggplot(cdf_data, aes(x = cum_pct, y = cost, color = claim_type)) +
  geom_line(linewidth = 1, alpha = 0.8) +
  scale_y_log10(
    labels = scales::comma,
    limits = c(1.0, NA) ) +
  labs(
    title = "Cumulative Distribution of Incurred Costs by Claim Type",
    x = "Cumulative Percentage of Claims (%)",
    y = "Incurred Cost (log scale)",
    color = "Claim Type"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 8),
    plot.title = element_text(face = "bold"),
    plot.margin = margin(10, 10, 30, 10)
  ) +
  guides(color = guide_legend(nrow = 3, ncol = 3, byrow = TRUE))
```

## The code

::: {style="font-size: 1.4em"}
```{r}
#| eval: false
# Reshape incurred cost data into long format
incurred_clean <- portfolio %>%
  select(
    liability_incurred,
    liability_property_incurred,
    liability_injury_incurred,
    property_incurred,
    theft_incurred,
    fire_incurred,
    glass_incurred,
    legal_protection_incurred,
    occupants_incurred
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "claim_type",
    values_to = "cost"
  ) %>%
  filter(is.finite(cost), cost > 0)

# ECDF transformation per claim type
cdf_data <- incurred_clean %>%
  group_by(claim_type) %>%
  arrange(cost) %>%
  mutate(
    cum_pct = row_number() / n() * 100
  ) %>%
  ungroup()

# Plot cumulative distribution
ggplot(cdf_data, aes(x = cum_pct, y = cost, color = claim_type)) +
  geom_line(linewidth = 1, alpha = 0.8) +
  scale_y_log10(
    labels = scales::comma,
    limits = c(1.0, NA) ) +
  labs(
    title = "Cumulative Distribution of Incurred Costs by Claim Type",
    x = "Cumulative Percentage of Claims (%)",
    y = "Incurred Cost (log scale)",
    color = "Claim Type"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 8),
    plot.title = element_text(face = "bold"),
    plot.margin = margin(10, 10, 30, 10)
  ) +
  guides(color = guide_legend(nrow = 3, ncol = 4, byrow = TRUE))
```
:::
::::

::: callout-note
## Insight:

The plot reveals that certain claim types, such as liability injury and fire claims, have much higher incurred costs at the upper end of the distribution compared to others. This indicates that while most claims are relatively small, a small proportion of catastrophic claims drive portfolio volatility and can heavily impact profitability.
:::

## 4.2 Premium Distribution

This code reshapes the premium variables into a long format and uses a stacked density plot on a logarithmic scale to visualize the distribution of premiums across different insurance types. The purpose is to compare how premiums are spread and concentrated among liability, property damage, theft, fire, glass, legal protection, and occupants coverage.

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
# Reshape incurred cost data into long format
incurred_long <- portfolio %>%
  select(
    liability_premium,
    property_damage_premium,
    theft_premium,
    fire_premium,
    glass_premium,
    legal_protection_premium,
    occupants_premium
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "insurance_type",
    values_to = "premium"
  )

# Stacked density plot
ggplot(incurred_long %>% filter(!is.na(premium), premium > 0),
       aes(x = premium, fill = insurance_type)) +
  geom_density(alpha = 0.6, position = "stack") +
  scale_x_log10(labels = comma) +   # log scale to handle skewness
  theme_minimal() +
  labs(
    title = "Stacked Density Plot of Premium",
    x = "Premuim (log scale)",
    y = "Density"
  ) +
  theme(
    strip.text = element_text(face = "bold"),
    legend.title = element_blank()
  )
```

## The code

::: {style="font-size: 1.4em"}
```{r}
#| eval: false
# Reshape incurred cost data into long format
incurred_long <- portfolio %>%
  select(
    liability_premium,
    property_damage_premium,
    theft_premium,
    fire_premium,
    glass_premium,
    legal_protection_premium,
    occupants_premium
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "insurance_type",
    values_to = "premium"
  )

# Stacked density plot
ggplot(incurred_long %>% filter(!is.na(premium), premium > 0),
       aes(x = premium, fill = insurance_type)) +
  geom_density(alpha = 0.6, position = "stack") +
  scale_x_log10(labels = comma) +   # log scale to handle skewness
  theme_minimal() +
  labs(
    title = "Stacked Density Plot of Premium",
    x = "Premuim (log scale)",
    y = "Density"
  ) +
  theme(
    strip.text = element_text(face = "bold"),
    legend.title = element_blank()
  )
```
:::
::::

::: callout-note
## Insight:

The plot shows that most premiums cluster at lower values, but the log scale reveals long tails where certain insurance types extend into higher premium ranges. Liability and property damage premiums dominate the distribution, while niche coverages like glass or legal protection appear less frequent and concentrated. This highlights that core coverages drive the bulk of premium income, whereas specialized coverages contribute smaller but distinct segments.
:::

## 4.3 Claim Distribution

This code reshapes the claim count variables into a long format and uses boxplots to visualize the distribution of claim numbers across different claim types. The purpose is to compare the frequency and spread of claims for liability, property, theft, fire, glass, legal protection, and occupants coverage.

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
# Reshape claim count columns into long format
claims_long <- portfolio %>%
  select(
    liability_claims,
    liability_property_claims,
    liability_injury_claims,
    property_claims,
    theft_claims,
    fire_claims,
    glass_claims,
    legal_protection_claims,
    occupants_claims
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "claim_type",
    values_to = "claim_count"
  ) %>%
  filter(is.finite(claim_count))

# Plot boxplots
ggplot(claims_long, aes(x = claim_type, y = claim_count, fill = claim_type)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.4) +
  labs(
    title = "Distribution of Claim Counts by Type",
    x = "Claim Type",
    y = "Number of Claims"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold")
  )
```

## The code

::: {style="font-size: 1.4em"}
```{r}
#| eval: false
# Reshape claim count columns into long format
claims_long <- portfolio %>%
  select(
    liability_claims,
    liability_property_claims,
    liability_injury_claims,
    property_claims,
    theft_claims,
    fire_claims,
    glass_claims,
    legal_protection_claims,
    occupants_claims
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "claim_type",
    values_to = "claim_count"
  ) %>%
  filter(is.finite(claim_count))

# Plot boxplots
ggplot(claims_long, aes(x = claim_type, y = claim_count, fill = claim_type)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.4) +
  labs(
    title = "Distribution of Claim Counts by Type",
    x = "Claim Type",
    y = "Number of Claims"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold")
  )
```
:::
::::

::: callout-note
## Insight:

The plot shows that most claim types have relatively low counts, with the majority clustering near zero, but property claims and liability‑related claims tend to have higher variability and more frequent occurrences. This indicates that property and liability claims are a major driver of claim activity, while categories like fire or glass are less common but still contribute to overall portfolio risk.
:::

## 4.4 Profitability Metrics

loss_ratio = total_incurred / total_premium\
→ Shows what fraction of premium income is consumed by claims. A ratio \> 1 means claims exceed premiums (unprofitable).

profit = total_premium - total_incurred\
→ The net financial result per policy. Positive = profitable, negative = loss.

This code categorizes policies into bins based on their loss ratio and profit values, then uses bar charts to visualize the distribution of policies across these categories. The purpose is to quickly assess how many policies are profitable, unprofitable, or break even, and to identify the proportion of policies with high or low loss ratios.

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
portfolio_bin <- portfolio %>%
  filter(has_premium) %>%
  mutate(
    # Loss ratio bins
    loss_ratio_bin = case_when(
      loss_ratio == 0 ~ "=0",
      loss_ratio > 0 & loss_ratio <= 1 ~ "0-1",
      loss_ratio > 1 ~ ">1",
      TRUE ~ NA_character_
    ),
    # Profit bins
    profit_bin = case_when(
      profit < 0 ~ "Negative",
      profit >= 0 & profit < 250 ~ "0-250",
      profit >= 250 & profit < 500 ~ "250-500",
      profit >= 500 ~ ">500",
      TRUE ~ NA_character_
    )
  )

p1 <- ggplot(portfolio_bin, aes(x = loss_ratio_bin, fill = loss_ratio_bin)) +
  geom_bar(alpha = 0.7) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  labs(title = "Loss Ratio Bins", x = "Loss Ratio Category", y = "Count") +
  scale_x_discrete(drop = TRUE) +
  theme_minimal() +
  theme(legend.position = "none")

p2 <- ggplot(portfolio_bin, aes(x = profit_bin, fill = profit_bin)) +
  geom_bar(alpha = 0.7) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  labs(title = "Profit Bins", x = "Profit Category", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")

p1 + p2
```

## The code

::: {style="font-size: 1.4em"}
```{r}
#| eval: false
portfolio_bin <- portfolio %>%
  filter(has_premium) %>%
  mutate(
    # Loss ratio bins
    loss_ratio_bin = case_when(
      loss_ratio == 0 ~ "=0",
      loss_ratio > 0 & loss_ratio <= 1 ~ "0-1",
      loss_ratio > 1 ~ ">1",
      TRUE ~ NA_character_
    ),
    # Profit bins
    profit_bin = case_when(
      profit < 0 ~ "Negative",
      profit >= 0 & profit < 250 ~ "0-250",
      profit >= 250 & profit < 500 ~ "250-500",
      profit >= 500 ~ ">500",
      TRUE ~ NA_character_
    )
  )

p1 <- ggplot(portfolio_bin, aes(x = loss_ratio_bin, fill = loss_ratio_bin)) +
  geom_bar(alpha = 0.7) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  labs(title = "Loss Ratio Bins", x = "Loss Ratio Category", y = "Count") +
  scale_x_discrete(drop = TRUE) +
  theme_minimal() +
  theme(legend.position = "none")

p2 <- ggplot(portfolio_bin, aes(x = profit_bin, fill = profit_bin)) +
  geom_bar(alpha = 0.7) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  labs(title = "Profit Bins", x = "Profit Category", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")

p1 + p2
```
:::
::::

::: callout-note
## Insight:

The charts show that a large majority of policies have a loss ratio of zero, meaning no claims were made, while a smaller but significant group has loss ratios greater than one, indicating unprofitable policies where claims exceeded premiums. On the profit side, most policies fall into moderate profit ranges (0–250 and 250–500), but there is also a notable share of highly profitable policies (\>500). This highlights that while many policies generate steady profits, a subset of high‑loss policies erodes overall portfolio performance.
:::

## 4.5 Categorical Variables

This code reshapes categorical variables into a long format, calculates the percentage distribution of each category, and uses bar charts to visualize them side by side. The purpose is to provide a clear overview of how policy types, statuses, business types, payment frequencies, bonus scores, fuel types, municipality types, and circulation areas are distributed across the portfolio.

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
# Categorical variables: bar charts
# Select categorical variables
cat_long <- portfolio %>%
  select(
    policy_type,
    policy_status,
    business_type,
    payment_frequency,
    bonus_score,
    fuel_type,
    municipality_type,
    circulation_area
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  )

# Create percentage table within each variable
plot_data <- cat_long %>%
  group_by(variable, value) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(variable) %>%
  mutate(
    percent = count / sum(count),
    value = fct_reorder(value, percent)
  )

# Plot
ggplot(plot_data,
       aes(x = percent,
           y = value)) +
  geom_col(fill = "steelblue") +
  facet_wrap(~variable,
             scales = "free_y") +
  scale_x_continuous(labels = percent_format()) +
  labs(
    title = "Distribution of Categorical Variables",
    x = "Percentage",
    y = NULL
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold"),
    axis.text.y = element_text(size = 9)
  )
```

## The code

::: {style="font-size: 1.4em"}
```{r}
#| eval: false
# Categorical variables: bar charts
# Select categorical variables
cat_long <- portfolio %>%
  select(
    policy_type,
    policy_status,
    business_type,
    payment_frequency,
    bonus_score,
    fuel_type,
    municipality_type,
    circulation_area
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  )

# Create percentage table within each variable
plot_data <- cat_long %>%
  group_by(variable, value) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(variable) %>%
  mutate(
    percent = count / sum(count),
    value = fct_reorder(value, percent)
  )

# Plot
ggplot(plot_data,
       aes(x = percent,
           y = value)) +
  geom_col(fill = "steelblue") +
  facet_wrap(~variable,
             scales = "free_y") +
  scale_x_continuous(labels = percent_format()) +
  labs(
    title = "Distribution of Categorical Variables",
    x = "Percentage",
    y = NULL
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold"),
    axis.text.y = element_text(size = 9)
  )
```
:::
::::

::: callout-note
## Insight:

The charts show that most policies are active and concentrated in portfolio business rather than new business, with annual payment frequency being the most common. Bonus scores lean heavily toward “Good,” while fuel types are dominated by gasoline, though a notable share is marked “Unknown.” Urban circulation areas and coastal municipalities are more prevalent, reflecting demographic and geographic concentration. These distributions highlight the portfolio’s structural composition and help explain underlying profitability and risk patterns.
:::

# 5 Exploratory Data Analysis

Apply suitable bivariate visual analytics techniques to investigate and statistically validate factors influencing the profitability and volatility of the insurance portfolio.

### 5.1 Correlation Analysis

cor() together with corrplot.mixed() selects key numeric variables, computes a Spearman correlation matrix, and visualizes it using a correlation plot with both ellipses and numeric values. It also extracts pairs with strong correlations and runs a statistical test between claim frequency and loss ratio. The purpose is to identify which financial, demographic, and risk metrics are most strongly related to profitability and volatility in the insurance portfolio.

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
# Select numeric variables
num_vars <- portfolio %>%
  select(total_premium, total_incurred, profit, loss_ratio,
         claim_frequency, claim_severity, total_claims,
         premium_per_exposure, incurred_per_exposure,
         driver_age, vehicle_age, power_to_weight_ratio, vehicle_value)

# Correlation matrix
cor_matrix <- cor(num_vars, use = "complete.obs", method = "spearman")

# Visualize
corrplot.mixed(cor_matrix,
               lower = "ellipse",
               upper = "number",
               number.cex = 0.5,
               tl.cex = 0.8,
               order = "hclust",
               tl.pos = "lt",
               diag = "l",
               tl.col = "black",
               na.label = " ")   

# Example correlation test
cor.test(portfolio$claim_frequency, portfolio$loss_ratio, method = "spearman")
```

## The code

::: {style="font-size: 1.4em"}
```{r}
#| eval: false
# Select numeric variables
num_vars <- portfolio %>%
  select(total_premium, total_incurred, profit, loss_ratio,
         claim_frequency, claim_severity, total_claims,
         premium_per_exposure, incurred_per_exposure,
         driver_age, vehicle_age, power_to_weight_ratio, vehicle_value)

# Correlation matrix
cor_matrix <- cor(num_vars, use = "complete.obs", method = "spearman")

# Visualize
corrplot.mixed(cor_matrix,
               lower = "ellipse",
               upper = "number",
               number.cex = 0.5,
               tl.cex = 0.8,
               order = "hclust",
               tl.pos = "lt",
               diag = "l",
               tl.col = "black",
               na.label = " ")   

# Example correlation test
cor.test(portfolio$claim_frequency, portfolio$loss_ratio, method = "spearman")
```
:::
::::

This code selects key numeric variables, computes a Spearman correlation matrix, and visualizes it using a correlation plot with both ellipses and numeric values. It also extracts pairs with strong correlations (\|ρ\| ≥ 0.5) and runs a statistical test between claim frequency and loss ratio. The purpose is to identify which financial, demographic, and risk metrics are most strongly related to profitability and volatility in the insurance portfolio.

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
cor_df <- cor_matrix %>%
  as.data.frame() %>%
  rownames_to_column("var1") %>%
  pivot_longer(-var1, names_to = "var2", values_to = "correlation") %>%
  filter(var1 < var2, abs(correlation) >= 0.5) %>%
  arrange(desc(abs(correlation)))

print(cor_df)
```

## The code

::: {style="font-size: 1.4em"}
```{r}
#| eval: false
cor_df <- cor_matrix %>%
  as.data.frame() %>%
  rownames_to_column("var1") %>%
  pivot_longer(-var1, names_to = "var2", values_to = "correlation") %>%
  filter(var1 < var2, abs(correlation) >= 0.5) %>%
  arrange(desc(abs(correlation)))

print(cor_df)
```
:::
::::

::: callout-note
## Insight:

Profitability is most negatively affected by high incurred costs and loss ratios (ρ ≈ –0.93 to –0.94), showing that claims relative to premiums directly erode profits.

Claim severity is strongly correlated with incurred costs and loss ratio (ρ ≈ 0.87–0.91), highlighting that catastrophic claims destabilize the portfolio more than claim frequency.

Claim frequency is highly correlated with loss ratio (ρ ≈ 0.90, p \< 2.2e‑16), confirming that frequent claims drive unprofitability.

Demographic factors like driver age and vehicle age show moderate correlation (ρ ≈ 0.86), but financial ratios remain the dominant drivers of portfolio performance.
:::

## 5.2 Profitability vs Risk Metrics

### 5.2.1 Premium vs Incurred

This code filters policies with positive incurred costs and premiums, then plots a scatter diagram with a regression line on a log scale. The purpose is to examine how incurred claim costs scale with premium levels and to assess whether higher premiums correspond to higher incurred losses.

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
# Premium vs Incurred (profitability)

portfolio_clean <- portfolio %>%
  filter(total_incurred > 0, total_premium >= 0)

ggplot(portfolio, aes(x = total_premium, y = total_incurred)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  scale_y_log10() +
  coord_cartesian(xlim = c(0, 3000), ylim = c(1, 500000)) +  
  geom_smooth(method = "lm", se = TRUE, color = "darkred") +
  labs(
    title = "Premium vs Incurred Costs",
    x = "Total Premium",
    y = "Total Incurred"
  ) +
  theme_minimal(base_size = 14)
```

## The code

::: {style="font-size: 1.4em"}
```{r}
#| eval: false
portfolio_clean <- portfolio %>%
  filter(total_incurred > 0, total_premium >= 0)

ggplot(portfolio, aes(x = total_premium, y = total_incurred)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  scale_y_log10() +
  coord_cartesian(xlim = c(0, 3000), ylim = c(1, 500000)) +  
  geom_smooth(method = "lm", se = TRUE, color = "darkred") +
  labs(
    title = "Premium vs Incurred Costs",
    x = "Total Premium",
    y = "Total Incurred"
  ) +
  theme_minimal(base_size = 14)
```
:::
::::

::: callout-note
## Insight:

There is a clear positive relationship between premiums and incurred costs, as shown by the upward regression line.

Most policies cluster at lower premium and incurred values, but variability increases significantly at higher premium levels.

The log scale highlights that while many policies have modest incurred costs, a small number of high‑premium policies are associated with very large claims.

This suggests that although higher premiums generate more income, they also expose the portfolio to greater volatility from large claims.
:::

### 5.2.2 Loss Ratio vs Claim frequency

This code filters policies with premiums, then plots a scatter diagram of claim frequency against loss ratio with a regression line. The purpose is to test whether higher claim frequency is associated with higher loss ratios, providing statistical validation of how frequent claims affect portfolio profitability.

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
ggplot(portfolio %>% filter(has_premium), aes(x = claim_frequency, y = loss_ratio)) +
  geom_point(alpha = 0.4, color = "steelblue") +   # scatter points
  coord_cartesian(xlim = c(0, 100), ylim = c(0, 1000)) +
  geom_smooth(method = "lm", se = TRUE, color = "darkred") +  # regression line
  labs(
    title = "Loss Ratio vs Claim Frequency",
    x = "Claim Frequency (claims per exposure)",
    y = "Loss Ratio"
  ) +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"))
```

## The code

::: {style="font-size: 1.4em"}
```{r}
#| eval: false
ggplot(portfolio  %>% filter(has_premium), aes(x = claim_frequency, y = loss_ratio)) +
  geom_point(alpha = 0.4, color = "steelblue") +   # scatter points
  coord_cartesian(xlim = c(0, 100), ylim = c(0, 1000)) +
  geom_smooth(method = "lm", se = TRUE, color = "darkred") +  # regression line
  labs(
    title = "Loss Ratio vs Claim Frequency",
    x = "Claim Frequency (claims per exposure)",
    y = "Loss Ratio"
  ) +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"))
```
:::
::::

::: callout-note
## Insight:

The upward regression line shows a strong positive relationship: as claim frequency increases, loss ratios also rise.

Most policies cluster at low claim frequency and low loss ratio, but a subset with very high claim frequency drives extreme loss ratios.

This confirms that frequent claims erode profitability, even when individual claim amounts are not catastrophic.

Managing claim frequency is therefore critical for stabilizing portfolio performance and reducing unprofitable policies.
:::

### 5.2.3 Profit vs Claim Severity

This code bins claim severity into ranges, calculates median profit for each bin, and visualizes the relationship using boxplots with annotated medians. The purpose is to show how increasing claim severity impacts profitability, highlighting the financial consequences of small versus catastrophic claims.

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
# Step 1: Create severity bins
portfolio_sev <- portfolio %>%
  mutate(
    severity_bin = case_when(
      claim_severity < 1000 ~ "<1k",
      claim_severity >= 1000 & claim_severity < 5000 ~ "1k-5k",
      claim_severity >= 5000 & claim_severity < 20000 ~ "5k-20k",
      claim_severity >= 20000 ~ ">20k",
      TRUE ~ NA_character_
    )
  )

# Step 2: Compute median profit per severity bin
summary_stats <- portfolio_sev %>%
  filter(!is.na(severity_bin)) %>%
  group_by(severity_bin) %>%
  summarise(median_profit = median(profit, na.rm = TRUE))

# Step 3: Boxplot with summary labels
ggplot(portfolio_sev %>% filter(!is.na(severity_bin)),
       aes(x = severity_bin, y = profit, fill = severity_bin)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  coord_cartesian(ylim = c(-400000 , 200000 )) +
  geom_text(data = summary_stats,
            aes(x = severity_bin, y = median_profit,
                label = scales::comma(round(median_profit, 0))),
            vjust = -0.5, fontface = "bold", color = "darkred") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Profit vs Claim Severity",
    x = "Claim Severity Bin",
    y = "Profit"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold")
  )
```

## The code

::: {style="font-size: 1.4em"}
```{r}
#| eval: false
# Step 1: Create severity bins
portfolio_sev <- portfolio %>%
  mutate(
    severity_bin = case_when(
      claim_severity < 1000 ~ "<1k",
      claim_severity >= 1000 & claim_severity < 5000 ~ "1k-5k",
      claim_severity >= 5000 & claim_severity < 20000 ~ "5k-20k",
      claim_severity >= 20000 ~ ">20k",
      TRUE ~ NA_character_
    )
  )

# Step 2: Compute median profit per severity bin
summary_stats <- portfolio_sev %>%
  filter(!is.na(severity_bin)) %>%
  group_by(severity_bin) %>%
  summarise(median_profit = median(profit, na.rm = TRUE))

# Step 3: Boxplot with summary labels
ggplot(portfolio_sev %>% filter(!is.na(severity_bin)),
       aes(x = severity_bin, y = profit, fill = severity_bin)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  coord_cartesian(ylim = c(-400000 , 200000 )) +
  geom_text(data = summary_stats,
            aes(x = severity_bin, y = median_profit,
                label = scales::comma(round(median_profit, 0))),
            vjust = -0.5, fontface = "bold", color = "darkred") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Profit vs Claim Severity",
    x = "Claim Severity Bin",
    y = "Profit"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold")
  )
```
:::
::::

::: callout-note
## Insight:

Policies with low claim severity (\<1k) generally remain profitable, with median profit close to zero but positive.

As claim severity increases (1k–5k and 5k–20k), median profits turn negative, showing that moderate claims already erode profitability.

The highest severity bin (\>20k) has the largest negative median profit (≈ –70k) and widest spread, confirming catastrophic claims are the biggest threat to portfolio stability.

Overall, profitability declines sharply as claim severity rises, emphasizing the importance of managing high‑cost claims to protect portfolio performance.
:::

## 5.3 Policy & Vehicle Characteristics

### 5.3.1 Policy & Profit

This code calculates the mean profit for each policy type, adds confidence intervals using bootstrapped error bars, and visualizes the results with a bar chart. The purpose is to compare average profitability across different policy types and assess which policies contribute most to portfolio performance.

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
# Mean profit per policy type with confidence intervals
ggplot(portfolio, aes(
    x = fct_reorder(policy_type, profit, .fun = mean, .desc = TRUE),
    y = profit,
    fill = policy_type
  )) +
  stat_summary(fun = mean, geom = "bar", alpha = 0.7) +
  stat_summary(fun.data = mean_cl_boot, geom = "errorbar", width = 0.2) +
  stat_summary(
    fun = mean,
    geom = "text",
    aes(label = scales::comma(..y..)),
    vjust = -0.5, size = 3.5, color = "black"
  ) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Mean Profit by Policy Type",
    x = "Policy Type",
    y = "Mean Profit"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold")
  )
```

## The code

::: {style="font-size: 1.4em"}
```{r}
#| eval: false
ggplot(portfolio, aes(
    x = fct_reorder(policy_type, profit, .fun = mean, .desc = TRUE),
    y = profit,
    fill = policy_type
  )) +
  stat_summary(fun = mean, geom = "bar", alpha = 0.7) +
  stat_summary(fun.data = mean_cl_boot, geom = "errorbar", width = 0.2) +
  stat_summary(
    fun = mean,
    geom = "text",
    aes(label = scales::comma(..y..)),
    vjust = -0.5, size = 3.5, color = "black"
  ) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Mean Profit by Policy Type",
    x = "Policy Type",
    y = "Mean Profit"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold")
  )
```
:::
::::

::: callout-note
## Insight:

COMP_E policies generate the highest mean profit, suggesting they are the most financially favorable product line.

COMP_N policies show the lowest mean profit, indicating weaker performance and potential risk to overall profitability.

Other policy types (CC, TP, TPG) fall in between, with moderate profitability compared to COMP_E.

The confidence intervals highlight variability, but the overall ranking of policy types remains consistent, guiding strategic focus toward more profitable products.
:::

### 5.3.2 Fuel Type vs Loss Ratio

This code compares loss ratio distributions across fuel types using violin plots overlaid with boxplots. The purpose is to assess whether different fuel types (Diesel, Gasoline, Unknown) are associated with distinct risk profiles and profitability outcomes.

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
# Violin plot: Loss ratio by fuel type
ggplot(portfolio %>% filter(has_premium), aes(x = fuel_type, y = loss_ratio, fill = fuel_type)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  coord_cartesian(ylim = c(0 , 5000)) +
  geom_boxplot(width = 0.1, fill = "white", outlier.alpha = 0.3) +
  labs(
    title = "Loss Ratio Distribution by Fuel Type",
    x = "Fuel Type",
    y = "Loss Ratio"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold")
  )
```

## The code

::: {style="font-size: 1.4em"}
```{r}
#| eval: false
ggplot(portfolio %>% filter(has_premium), aes(x = fuel_type, y = loss_ratio, fill = fuel_type)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  coord_cartesian(ylim = c(0 , 5000)) +
  geom_boxplot(width = 0.1, fill = "white", outlier.alpha = 0.3) +
  labs(
    title = "Loss Ratio Distribution by Fuel Type",
    x = "Fuel Type",
    y = "Loss Ratio"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold")
  )
```
:::
::::

::: callout-note
## Insight:

Both Diesel and Gasoline vehicles show wide variability in loss ratios, with several extreme outliers exceeding 3000, indicating higher volatility and risk exposure.

Median loss ratios are relatively similar across categories, but the spread for Diesel and Gasoline highlights that these segments carry disproportionate risk from outlier claims.

This suggests that fuel type may be a useful segmentation variable for underwriting, with Diesel and Gasoline requiring closer monitoring for high‑loss policies.
:::

## 5.4 Driver Age vs Claim Severity

This compares the distribution of loss ratios across age bands. This highlights which age groups tend to have higher or lower loss ratios.

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
portfolio_age <- portfolio %>% 
  mutate(
    age_band = cut(driver_age,
                   breaks = seq(18, 80, by = 10),
                   include.lowest = TRUE)
  ) %>%
  filter(!is.na(age_band)) %>%   # <-- remove NA age bands
  group_by(age_band) %>%
  mutate(
    p25   = quantile(claim_severity, 0.25, na.rm = TRUE),
    q3    = quantile(claim_severity, 0.75, na.rm = TRUE),
    iqr   = q3 - p25,
    lower = p25,                  
    upper = q3 + 1.5 * iqr        
  ) %>%
  ungroup() %>%
  filter(claim_severity >= lower, claim_severity <= upper)

# Ridge plot of loss ratio by age band (log scale, NA removed)
ggplot(portfolio_age, aes(x = loss_ratio, y = age_band, fill = age_band)) +
  ggridges::geom_density_ridges(alpha = 0.6) +
  scale_x_log10() +
  labs(
    title = "Claim Severity Distribution by Driver Age Band (P25 cutoff, log scale)",
    x = "Log Claim Severity",
    y = "Driver Age Band"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```

## The code

::: {style="font-size: 1.4em"}
```{r}
#| eval: false
# Create age bands
portfolio_age <- portfolio %>% 
  mutate(
    age_band = cut(driver_age,
                   breaks = seq(18, 80, by = 10),
                   include.lowest = TRUE)
  ) %>%
  filter(!is.na(age_band)) %>%   # <-- remove NA age bands
  group_by(age_band) %>%
  mutate(
    p25   = quantile(claim_severity, 0.25, na.rm = TRUE),
    q3    = quantile(claim_severity, 0.75, na.rm = TRUE),
    iqr   = q3 - p25,
    lower = p25,                  
    upper = q3 + 1.5 * iqr        
  ) %>%
  ungroup() %>%
  filter(claim_severity >= lower, claim_severity <= upper)

# Ridge plot of loss ratio by age band (log scale, NA removed)
ggplot(portfolio_age, aes(x = loss_ratio, y = age_band, fill = age_band)) +
  ggridges::geom_density_ridges(alpha = 0.6) +
  scale_x_log10() +
  labs(
    title = "Claim Severity Distribution by Driver Age Band (P25 cutoff, log scale)",
    x = "Log Claim Severity",
    y = "Driver Age Band"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```
:::
::::

::: callout-note
## Insight:
Claim severity patterns vary across age bands, but differences are modest once extreme outliers are removed.

Younger drivers (18–28) show slightly higher severity risk.

Middle‑aged drivers (38–58) are more stable and predictable.

Older drivers (68–78) again show wider variability.

Overall, claim severity is most stable in mid‑age groups, while both younger and older drivers add more volatility.
:::

## 5.5 Volatility

To measure how stable or unstable underwriting results are by analyzing claim frequency, claim severity, exposure relationships, and variance metrics. This helps identify which segments of the portfolio contribute most to volatility.

### 5.5.1 Loss ratio variability

The code chunk uses summarise() to calculate the mean, standard deviation, and variance of the loss ratio by policy type. This measures how much the ratio of incurred to premium fluctuates across policies, where high variance indicates unstable underwriting results.

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
# Variance of loss ratio by policy type
portfolio %>%
  group_by(policy_type) %>%
  summarise(
    mean_lr = mean(loss_ratio, na.rm = TRUE),
    sd_lr   = sd(loss_ratio, na.rm = TRUE),
    var_lr  = var(loss_ratio, na.rm = TRUE)
  )
```

## The code

::: {style="font-size: 1.4em"}
```{r}
#| eval: false
portfolio %>%
  group_by(policy_type) %>%
  summarise(
    mean_lr = mean(loss_ratio, na.rm = TRUE),
    sd_lr   = sd(loss_ratio, na.rm = TRUE),
    var_lr  = var(loss_ratio, na.rm = TRUE)
  )
```
:::
::::

::: callout-note
## Insight:

Loss ratio variance highlights how much incurred/premium fluctuates across policy types.

COMP_E shows the highest variance (≈1520), meaning highly unstable results.

COMP_N has the lowest variance (≈18), indicating more predictable performance.

High variance signals segments where underwriting outcomes are less reliable.
:::

### 5.5.2 Coefficient of Variation

CV normalizes volatility relative to the mean, making it easier to compare stability across different measures.

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
# CV for claim frequency and severity
portfolio %>%
  summarise(
    mean_freq = mean(claim_frequency, na.rm = TRUE),
    sd_freq   = sd(claim_frequency, na.rm = TRUE),
    cv_freq   = sd_freq / mean_freq,
    
    mean_sev = mean(claim_severity, na.rm = TRUE),
    sd_sev   = sd(claim_severity, na.rm = TRUE),
    cv_sev   = sd_sev / mean_sev
  )
```

## The code

::: {style="font-size: 1.4em"}
```{r}
#| eval: false
portfolio %>%
  summarise(
    mean_freq = mean(claim_frequency, na.rm = TRUE),
    sd_freq   = sd(claim_frequency, na.rm = TRUE),
    cv_freq   = sd_freq / mean_freq,
    
    mean_sev = mean(claim_severity, na.rm = TRUE),
    sd_sev   = sd(claim_severity, na.rm = TRUE),
    cv_sev   = sd_sev / mean_sev
  )
```
:::
::::

::: callout-note
## Insight:

CV normalizes volatility relative to the mean.

Claim frequency CV ≈ 6.17 → extremely unstable relative to its average.

Claim severity CV ≈ 3.45 → also volatile, but less extreme than frequency.

Frequency volatility dominates, showing instability in how often claims occur.
:::

### 5.5.3 Levene's Test

The code chunk applies leveneTest() to check whether the variance of loss ratio and claim severity differs significantly across groups (circulation area and fuel type). This statistically validates whether volatility is group‑dependent.

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
library(car)

# Levene’s test for loss ratio variance across circulation areas
leveneTest(loss_ratio ~ circulation_area, data = portfolio)

# Levene’s test for claim severity variance across fuel types
leveneTest(claim_severity ~ fuel_type, data = portfolio)
```

## The code

::: {style="font-size: 1.4em"}
```{r}
#| eval: false
# Levene’s test for loss ratio variance across circulation areas
leveneTest(loss_ratio ~ circulation_area, data = portfolio)

# Levene’s test for claim severity variance across fuel types
leveneTest(claim_severity ~ fuel_type, data = portfolio)
```
:::
::::

::: callout-note
## Insight:

Tests whether variance differs across groups (e.g., urban vs rural, fuel types).

Results show no significant differences (p \> 0.05), meaning variance is statistically similar across categories.

Suggests volatility is driven more by claim behavior than by geography or fuel type.
:::

### 5.5.4 Time-series Volatility

The code chunk calculates rolling standard deviation of the incurred/premium ratio over time using rollapply(). This tracks volatility dynamically, showing how portfolio stability changes across years.

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
library(zoo)
portfolio_ts <- portfolio %>%
  group_by(year) %>%
  summarise(ratio = mean(total_incurred / total_premium, na.rm = TRUE))

portfolio_ts$rolling_sd <- rollapply(portfolio_ts$ratio, width = 3, FUN = sd, fill = NA, align = "right")
```

## The code

::: {style="font-size: 1.4em"}
```{r}
#| eval: false
portfolio_ts <- portfolio %>%
  group_by(year) %>%
  summarise(ratio = mean(total_incurred / total_premium, na.rm = TRUE))

portfolio_ts$rolling_sd <- rollapply(portfolio_ts$ratio, width = 3, FUN = sd, fill = NA, align = "right")
```
:::
::::

::: callout-note
## Insight:

Rolling standard deviation of loss ratios shows periods of instability over time.

Peaks indicate claim shocks or seasonal effects.

Confirms that volatility is not constant — portfolio risk fluctuates across years.
:::

## 5.6 Hypothesis Testing

To statistically validate whether differences in profitability, claims, or loss ratios across categories are significant. This ensures observed patterns are not due to random variation.

### 5.6.1 Chi-square tests

The code chunk runs chisq.test() on contingency tables to test independence between categorical variables (policy type, bonus score, circulation area) and outcomes (profitability, claim occurrence). This validates whether distributions differ significantly across categories.

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
chisq.test(table(portfolio$policy_type, portfolio$profitable_policy))
chisq.test(table(portfolio$bonus_score, portfolio$profitable_policy))
chisq.test(table(portfolio$circulation_area, portfolio$has_claim))
```

## The code

::: {style="font-size: 1.4em"}
```{r}
#| eval: false
chisq.test(table(portfolio$policy_type, portfolio$profitable_policy))
chisq.test(table(portfolio$bonus_score, portfolio$profitable_policy))
chisq.test(table(portfolio$circulation_area, portfolio$has_claim))
```
:::
::::

::: callout-note
## Insight:

Policy type vs profitability: Highly significant (p \< 2.2e‑16) → profitability depends on policy type.

Bonus score vs profitability: Significant → Good/Neutral/Bad scores matter for outcomes.

Circulation area vs has_claim: Significant → urban vs rural differ in claim occurrence.
:::

### 5.6.2 ANOVA

The code chunk applies aov() and TukeyHSD() to compare mean loss ratios across fuel types. This tests whether continuous outcomes differ significantly between categorical groups.

:::: panel-tabset
## The plot

```{r}
#| echo: false
#｜ fig-width: 12
#｜ fig_height: 8
anova_result <- aov(loss_ratio ~ fuel_type, data = portfolio)
summary(anova_result)
TukeyHSD(anova_result)
```

## The code

::: {style="font-size: 1.4em"}
```{r}
#| eval: false
anova_result <- aov(loss_ratio ~ fuel_type, data = portfolio)
summary(anova_result)
TukeyHSD(anova_result)
```
:::
::::

::: callout-note
## Insight:

Loss ratio vs fuel type: Not significant (p = 0.38) → no mean difference across Diesel/Gasoline/Unknown.

Profit vs business type: Useful for comparing New Business vs Portfolio profitability.

Claim severity vs municipality type: ANOVA compares Inland vs Coastal vs Islands; differences highlight geographic claim risk.
:::

# 6 Key Insight Summary

**Claims behavior is the dominant source of volatility.**

-   Claim frequency is highly unstable relative to its mean (CV ≈ 6.17), while claim severity also contributes (CV ≈ 3.45).

-   Catastrophic claims drive extreme incurred costs, while frequent smaller claims steadily erode profitability.

**Policy type and bonus score are critical profitability differentiators.**

-   Chi‑square tests confirm that profitability outcomes vary significantly by policy type and bonus score.

-   COMP_E policies show high variance and unstable underwriting results, while COMP_N policies are more predictable.

**Geography and fuel type play a limited role.**

-   Levene’s tests and ANOVA results show no significant differences in variance or mean loss ratios across circulation areas or fuel types.

-   This suggests that underwriting focus should prioritize claims behavior and product design over these categorical splits.

**Portfolio risk is dynamic over time.**

-   Rolling time‑series volatility reveals spikes linked to shocks or seasonal effects, confirming that risk is not constant.

-   Monitoring volatility trends is essential for proactive risk management.

**Visual analytics adds measurable value.**

-   Univariate plots highlight skewness and outliers.

-   Bivariate scatterplots and boxplots reveal correlations between premiums, incurred costs, claim frequency, and profitability.

-   Statistical tests validate which categorical differences are meaningful, ensuring decisions are evidence‑based.

# Overall Conclusion

The investigation shows that **profitability and volatility in motor insurance portfolios are primarily shaped by claims frequency and severity, with policy type and bonus score acting as key structural differentiators.** While geography and fuel type do not significantly explain volatility, portfolio risk is highly dynamic over time, requiring continuous monitoring.

By combining visual analytics with statistical validation, insurers can move beyond descriptive reporting to actionable insights:

-   Targeting product types with unstable underwriting results.

-   Designing interventions to manage claim frequency and severity.

-   Using bonus scores as a lever for risk segmentation.

-   Tracking volatility trends to anticipate shocks.

This demonstrates the potential of visual analytics not only to explore data distributions but also to identify the true drivers of profitability and volatility, ultimately guiding more stable underwriting and strategic portfolio management.
